---
title: "Cryptography and Coding Theory"
author: "David J. Hunter"
output:
  html_document:
    toc: true
    toc_float:
      collapsed: false
      smooth_scroll: true
    df_print: paged
    theme: spacelab
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, echo=FALSE}
library(metathis)
meta() %>%
  meta_description(
    "Slides, assignments, and code for my cryptography and coding theory class."
  ) %>%
  meta_name("github-repo" = "djhunter/cryptography") %>%
  meta_viewport() %>%
  meta_social(
    title = "Cryptography and Coding Theory",
    url = "https://djhunter.github.io/cryptography/index.html",
    image = "https://upload.wikimedia.org/wikipedia/en/5/5b/Codes_And_Keys_Death_Cab_For_Cutie.jpg",
    image_alt = "Codes and Keys album art",
    og_author = c("David J. Hunter")
  )
```

This site provides slides, assignments, and other resources that I use in my Codes and Encryption class (MA/CS-124, Westmont College, 2016-2020).

# Slides 

1. [Introduction: Cryptography and Coding Theory](slides/01intro_crypto.html) 
2. [Affine Ciphers and the GCD](slides/02affine_ciphers.html) 
3. [Substitution and Vigenère Ciphers](slides/03vigenere_ciphers.html)
4. [Attacks on Vigenère Ciphers](slides/04vigenere_attacks.html)
5. [Matrices and the Hill Cipher](slides/05matrix_hill.html)
6. [Linear Feedback Shift Registers](slides/06lfsr.html)
7. [Data Encryption Standard (DES); Finite Fields](slides/07desaes1.html)
8. [Advanced Encryption Standard (AES)](slides/08desaes2.html)
9. [RSA Public Key Encryption](slides/09rsa1.html)
10. [Attacks on RSA and Primality Testing](slides/10rsa2.html)
11. [Discrete Logarithms, Diffie-Hellman, ElGamal](slides/11discrete_logs.html)
12. [Hash Functions](slides/12hash_functions.html)
13. [Digital Signatures](slides/13digital_sigs.html)
14. [Real Elliptic Curves](slides/14elliptic_curves1.html)
15. [Mod p Elliptic Curves](slides/15elliptic_curves2.html)
16. [Elliptic Curve Cryptography](slides/16elliptic_curves3.html)
17. [Introduction to Codes](slides/17codes1.html)
18. [Codes: Definitions and Theorems](slides/18codes2.html)
19. [Linear Codes](slides/19codes3.html)
20. [Cyclic Codes](slides/20codes4.html)
21. [Fields and BCH Codes](slides/21codes5.html)
22. [Reed Solomon Codes](slides/22codes6.html)

# Written Assignments

1. [Modular arithmetic](assignments/01modding.html)
2. [Shift ciphers and affine ciphers](assignments/02shiftaffine.html)
3. [Substitution and Vigenère Ciphers](assignments/03subciphers.html)
4. [Properties of Vigenère Ciphers](assignments/04vigenere.html)
5. [Matrices and the Hill Cipher](assignments/05mathill.html)
6. [Matrices and LFSR's](assignments/06matlfsr.html)
7. [Meet-in-the-middle; Finite Fields](assignments/07mitmff.html)
8. [AES and Modes of Operation](assignments/08aesmodes.html)
9. [Some RSA Calculations](assignments/09rsacalc.html)
10. [Prime Numbers and RSA](assignments/10primes.html)
11. [Discrete Logarithms](assignments/11dlogs.html)
12. [Hash Functions and the Birthday Theorem](assignments/12hash.html)
13. [Digital Signatures](assignments/13sigs.html)
14. [Real Elliptic Curves](assignments/14real_ell.html)
15. [Mod p Elliptic Curves](assignments/15modp_ell.html)
16. [Elliptic Curve Cryptography](assignments/16ec_crypto.html)
17. [Codes: Introductory Examples](assignments/17introcodes.html)
18. [Codes: Theoretical Limitations](assignments/18codelimits.html)
19. [Systematic Codes](assignments/19systematic.html)
20. [Cyclic Codes](assignments/20cyclic.html)
21. [BCH Codes](assignments/21bch.html)
22. [Reed-Solomon Codes](assignments/22reedsolomon.html)

# Programming Assignments

1. [Getting started with R packages](programming/01affine.html)
2. [Tools for Breaking a Vigenère Cipher](programming/02vigbreak.html)
3. [OTP and LFSR tools](programming/03mathillotp.html)
4. [GF(256) and AES](programming/04gf256aes.html)
5. [RSA tools](programming/05rsa.html)
6. [Discrete Logarithms and Hash Functions](programming/06discloghash.html)
7. [DSA and Real Elliptic Curves](programming/07dsa_real_ell.html)
8. [Mod p Elliptic Curves](programming/08modp_ell.html)
9. [Code Distance and Generation](programming/09code_tools1.html)
10. [Systematic Codes](programming/10code_tools2.html)
11. [Matrix Tools for Codes](programming/11code_tools3.html)

# Other Resources

- [Syllabus](tex/macs124f20.pdf) (PDF)

All of the pages and slides on this site were authored in RMarkdown. The source code is available on GitHub: https://github.com/djhunter/cryptography

