---
title: "Introduction to Cryptography"
date: "August 31, 2020"
output: 
  revealjs::revealjs_presentation:
    fig_width: 14
    fig_height: 7
    self_contained: true
    theme: night
    highlight: zenburn
    css: slidesdjh.css
    center: false
    transition: slide
    reveal_options:
      controls: true
      progress: false
      width: 1080
      height: 540
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(comment = NA)
```

# Introduction to Cryptography

## What is Cryptography?

A sender (traditionally called Alice) would like to send a message to a receiver (traditionally, Bob). 

However, while the message is being sent, it can be observed by an eavesdropper (Eve), and Bob and Alice would like to hide it's meaning from Eve.

## Alice, Bob, Eve {.build}

For example, Alice wants to send the following message:

```
mysnapchatusernameisramblingscrambler
```

Instead, she sends the message:
```
oaupcrejcvwugtpcogkutcodnkpiuetcodngt
```

>  **Discuss.** Eve observes this communication, and figures out their encryption scheme. The next message sent is `tqhnqn`. What is the decrypted message? What is Alice and Bob's scheme?

## ~~What you need~~ What you will learn

- cleverness
- algorithms
- mathematics
- R

## What is it good for?

- Privacy
- E-Commerce
- Digital rights management
- Military/Espionage
- Crypto-currencies
- Authentication
- Certification
- Data integrity

## Is cryptography different from coding theory?

> **Coding Theory** deals with how we represent information digitally. 

- Error-correcting codes
- Compression
- Data transmission
- Cryptographic coding (i.e., cryptography)

[Syllabus](../tex/macs124f20.pdf) (PDF)

# Assumptions and Terminology 

## Kerckhoff's Principle 

> **Assumption:** Eve always knows the method Alice and Bob are using.

## Terminology 

>- **Plaintext:** The message that Alice would like to send to Bob. (`roflol`)

>- **Ciphertext:** The encrypted message that Alice sends (`tqhnqn`)

>- **Key:** A piece of information, which, combined with the method, allows a ciphertext to be decrypted.

---

> *Example.* The method described above is called a **shift cipher**
> (or Caesar cipher). 

- Plaintext: `roflol`
- Ciphertext: `tqhnqn`
- Key: `2` (how much to shift)

## The modulo equivalence relation 

We write 
$$a \equiv b \pmod n$$ to denote that $a$ and $b$ belong to the same equivalence class modulo $n$; that is, $(a-b)$ is a multiple of $n$. 

## The modulo operator 

We write $n \bmod m$ to represent the remainder when $n$ is divided by $m$. For example, in R, we compute $38 \bmod 6$ as follows.

```{r modop, echo=TRUE}
38 %% 6
```

Since $38 \bmod 6 = 2$, $38$ and $2$ are in the same equivalence class modulo $6$, which we denote as 
$$38 \equiv 2 \pmod 6$$
or equivalently, as
$$2 \equiv 38 \pmod 6$$

## Modular arithmetic

> **Theorem.** If $a \equiv b \pmod n$ and $c \equiv d \pmod n$, then 
> 
> - $a+c \equiv b+d \pmod n$
> - $ac \equiv bd \pmod n$

In other words, when you do arithmetic modulo $n$, it doesn't matter which equivalence class representatives you use. In terms of operators, you can apply the $\bmod$ operator before or after adding and multiplying, and you will obtain the same result. (Example.)

## Group Exercise: Modular arithmetic

> Compute each expression modulo 7, without using technology.
>
> 1. $(32 + 76)(144 - 7000) \bmod 7$
>
> #. $23^{100} \bmod 7$

## Modular arithmetic in R

What goes wrong here? 

```{r, echo=TRUE}
23^100 %% 7 # wrong!
```

## Big integer modular arthmetic in R

```{r, results='hide', message=FALSE, warning=FALSE, echo=TRUE}
library(gmp) # load library for multiple precision arithmetic
```

```{r, echo=TRUE}
a <- as.bigz(2)^100
a
a %% 7
```

----

Or better yet:

```{r, echo=TRUE}
powm(23, 100, 7)
```



## Shift cipher implemented in R

```{r shift, echo=TRUE}
stringToMod26 <- function(x) {utf8ToInt(x)-utf8ToInt("a")}
mod26ToString <- function(x) {intToUtf8(x+utf8ToInt("a"))}

shiftCipher <- function(p,b)
{
  pt <- stringToMod26(p)
  ct <- (pt + b) %% 26      # add b to each letter
  return(mod26ToString(ct))
}

shiftCipher("thisisasecretmessage", 2)
```

## Decrypting a shift cipher {.build}

```{r, echo=TRUE}
shiftCipher("cubswin", 10)
```

```{r, echo=TRUE}
shiftCipher("melcgsx", -10)
```

```{r, echo=TRUE}
shiftCipher("melcgsx", 16)
```

## Shift cipher: ciphertext only attack

> In a **ciphertext only attack**, Eve has only the ciphertext and wishes to decrypt it and obtain the key.

How could she decrypt it?

## Loops in R

```{r, echo=TRUE}
ciphertext <- "ufekwfixvkkfufkyvrjjzxedvekj"
for(i in 1:26) {
  print(shiftCipher(ciphertext, i))}
```

## The R way: use `sapply`
```{r, echo=TRUE}
sapply(1:26, function(x) {shiftCipher(ciphertext, x)})
```

## Shift cipher: known plaintext attack {.build}

> In a **known plaintext attack**, Eve has a plaintext string along with the corresponding ciphertext, and wishes to recover the key.

Ideas?

```{r, echo=TRUE}
plaintext <- "usecanvas"
ciphertext <- "camkivdia"
(stringToMod26(ciphertext)-stringToMod26(plaintext)) %% 26
```

## Shift cipher: chosen plaintext attack {.build}

> In a **chosen plaintext attack**, Eve temporarily gets access to the "encryption machine", and encrypts a plaintext message of her choosing, observing the ciphertext.

```{r, echo=TRUE}
key <- 7
stringToMod26(shiftCipher("a", key)) %% 26
```

## Shift cipher: chosen ciphertext attack {.build}

> In a **chosen ciphertext attack**, Eve temporarily gets access to the "decryption machine", and decrypts a ciphertext message of her choosing, observing the plaintext.

```{r, echo=TRUE}
key <- 7
-stringToMod26(shiftCipher("a", -key)) %% 26
```

## Assignment: due Wednesday, 11:59 pm.

 0. Install RStudio on your favorite computer. Install the library `gmp` and test some of the examples from these notes.

 1. Find the last digit of $2^{10000}$ (i.e., the one's digit). Use any method. Explain how you found your answer.


 
 #. Which of the following equations have *integer* solutions? For each equation, find an integer solution or explain why no such solution exists.
    a. $2x + 3 \equiv 1 \pmod 9$
    
    b. $2x + 3 \equiv 1 \pmod 8$
    
    c. $2x + 3 \equiv 2 \pmod 8$
    
    
 <!-- #. Suppose that an [official English Scrabble word](http://scrabble.hasbro.com/en-us/tools#dictionary) was encrypted using a shift cipher resulting in the ciphertext `vaxxk`. What are the possible keys? (Include R code and output that you use to answer this question.) -->

<br><br><br><br><br><br>


