---
title: "Affine Ciphers"
date: "September 3, 2020"
output: 
  revealjs::revealjs_presentation:
    fig_width: 14
    fig_height: 7
    self_contained: true
    theme: night
    highlight: zenburn
    css: slidesdjh.css
    center: false
    transition: slide
    reveal_options:
      controls: true
      progress: false
      width: 1080
      height: 540
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(comment = NA)
library(gmp)
```

# Last time: the Shift Cipher

## Ciphers

> A **cipher** is an algorithm that inputs a plaintext string and outputs a ciphertext string, according to some given key.

Example: The shift cipher.

Ciphers, as opposed to **codes**, operate on strings without regard to their meaning (semantics). 

## Recall the Shift Cipher

```{r shift, echo=TRUE}
stringToMod26 <- function(x) {utf8ToInt(x)-utf8ToInt("a")}
mod26ToString <- function(x) {intToUtf8(x+utf8ToInt("a"))}

shiftCipher <- function(p,b)
{
  pt <- stringToMod26(p)
  ct <- (pt + b) %% 26      # add b to each letter
  return(mod26ToString(ct))
}

shiftCipher("thisisasecretmessage", 2)
```

## Notation/Terminology

When we want to consider just the numbers $0, 1, 2, \ldots, n-1$, and we are doing addition and multiplication modulo $n$, we say that we are "working in $\mathbb{Z}_n$," pronounced "zee mod en". 

More formally, $\mathbb{Z}_n$ is the set $\{0, 1, 2, \ldots, n-1\}$ along with multiplication and addition modulo $n$. Such a set is called a **ring**. 

- In a ring, you can add and subtract and there's a zero.
- In a ring, you can multiply, but maybe not always divide.
- Examples: real numbers, square matrices, $\mathbb{Z}_n$.

## The Shift Cipher as a function

The shift cipher can be represented by a function $$s: \mathbb{Z}_{26} \longrightarrow \mathbb{Z}_{26}$$ given by 
$$x \stackrel{s}{\longmapsto} x+k$$
where $k \in \mathbb{Z}_{26}$ is the key. To implement the shift cipher, we had to translate strings to $\mathbb{Z}_{26}$ and back, but the processes of encryption and decryption are performed by the function $s$ and its inverse $s^{-1}$, respectively.

# The Affine Cipher

## Affine

In mathematics, an **affine** function/transformation performs both a shift and a scale. In $\mathbb{Z}_n$, shifting is done by adding (like the shift cipher), and scaling is done by multiplying. So an **affine cipher** corresponds to a function 
$$s: \mathbb{Z}_{26} \longrightarrow \mathbb{Z}_{26}$$ given by 
$$x \stackrel{s}{\longmapsto} \alpha x + \beta$$
where the key consist of two elements $\alpha, \beta \in \mathbb{Z}_{26}$.

## Groups for today

Join the *voice channel* for your group. Use the *Jamboard* (link on `#macs-124`).

```{r, echo=FALSE}
library(knitr)
roster <- c("Sam", "Wesley", "Andrew", "Winston", "Michael", "Logan", "Christian", "Caleb", "James", "Joe", "Emily", "Giselle", "Nathan")
set.seed(932020) 
n <- length(roster)
ngps <- 3
maxingp <- ceiling(n/ngps)
# just make random groups
groups <- matrix(c(roster[sample(n)], 
                   rep("",(maxingp - (n %% maxingp)) %% maxingp)), 
                 ncol=maxingp, byrow=FALSE)
rownames(groups) <- paste0("Group #", 1:nrow(groups))
kable(t(groups))
```

## Group Activity (without computers)

Consider the affine cipher given by $x \longmapsto 3x + 7$ in $\mathbb{Z}_{26}$. If `cat` is the plaintext, what is the ciphertext? If `dog` is the ciphertext, what is the plaintext?

```{r, echo=FALSE}
library(knitr)
ln <- t(as.matrix(0:25))
colnames(ln) <- letters
kable(ln)
```


```{r}
stringToMod26("cat")
stringToMod26("dog")
```

## Inverses

In $\mathbb{Z}_{n}$, we say that $b$ is a **multiplicative inverse** of $a$ if $ab = 1$. 

What is the multiplicative inverse of $3$ in $\mathbb{Z}_{26}$? Can we use this to calculate $s^{-1}$ in the above example?
Consider:

```{r}
0:25 
(3 * (0:25)) %% 26
```

## Inverses?

What is the multiplicative inverse of $4$ in $\mathbb{Z}_{26}$? 

```{r}
0:25 
(4 * (0:25)) %% 26
```

## Zero divisors are not invertible

- We call $4$ a *zero divisor* in the ring $\mathbb{Z}_n$ because $4 \cdot 13 = 0$ in this ring.
- A zero divisor $a$ is an element such that $ab = 0$ for some nonzero $b$.
- If $a$ is a zero divisor, then it can't be invertible:
    - **Proof:** Suppose to the contrary that $a$ is a zero divisor and $a^{-1}$ exists. Then $ab = 0$ for some nonzero $b$. Multiplying both sides of this equation by $a^{-1}$ gives $a^{-1}ab = a^{-1}0$, which simplifies to $b = 0$, contradicting that $b$ is nonzero.

# The GCD

## Greatest Common Divisor 

- **Definition.** The *greatest common divisor* $\gcd(m,n)$ of two integers $m$ and $n$ is the largest positive integer $g$ such that $g \mid m$ and $g \mid n$. The greatest common divisor $\gcd(m,n)$ has the property that, if $r$ is a common divisor of both $m$ and $n$, then $r \mid \gcd(m,n)$.
- **Theorem.** The element $a$ has a multiplicative inverse in $\mathbb{Z}_{n}$ if and only if $\gcd(a,n) = 1$.

We say that $a$ and $n$ are "relatively prime" if $\gcd(a,n)=1$.

## Euclidean Algorithm

> **Theorem.** For any natural numbers $a$ and $b$, there exist natural numbers $c$ and $d$ such that $\gcd(a,b)=ca + db$. 

This theorem is proved by defining an algorithm to find the GCD. This algorithm is called the *Euclidean algorithm*, and when it is modified to also give the numbers $c$ and $d$, it is called the *extended Euclidean algorithm.*

## Euclidean Algorithm: Example

$$
\begin{aligned}
1180 &= 2 \cdot 482 + 216 \\
482 &= 2\cdot 216 + 50 \\
216 &= 4\cdot 50 +16 \\
50 &= 3\cdot 16 + 2 \\
16 &= 8\cdot 2 + 0
\end{aligned}
$$

Therefore, $\gcd(1180,482)=2$, the last nonzero remainder. (why?)

## Extended Euclidean Algorithm  

We can find numbers $c$ and $d$ such that $1180c + 482d = 2$ by manipulating these equations. Starting with the second-to-last equation,

$$
{\small 
\begin{aligned}
2 &= 50-3\cdot 16 \\
&= 50-3(216-4\cdot 50) \\
&= 482-2\cdot 216 -3(216 - 4(482-2\cdot 216)) \\
&= 482-2(1180-2\cdot 482)-3((1180-2\cdot 482)-4(482-2(1180-2\cdot 482))) \\
&= (-2-3-24)1180 + (1+4+6+12+48)482 \\
&= - 29\cdot 1180 + 71 \cdot 482 
\end{aligned} }
$$


## Implementation of the extended Euclidean Algorithm in R 

```{r gmp, include = FALSE}
library(gmp)
```

```{r eucalglist, echo=TRUE}
extGCDbigz <- function (a, b, showTrace = FALSE) 
{
  u <- c(as.bigz(1), as.bigz(0), as.bigz(a))
  v <- c(as.bigz(0), as.bigz(1), as.bigz(b))
  if(showTrace)
      cat(paste("v =", v[1], v[2], v[3], "\n"))
    while (v[3] != 0) {
      # Invariant: v[1]*a + v[2]*b = v[3]
      q <- u[3] %/% v[3]
      t <- u - v * q
      u <- v
      v <- t
      if(showTrace)
        cat(paste("v =", v[1], v[2], v[3], "\n"))
    }
    return(list(g = u[3], c = u[1], d = u[2]))
}
```

## Implementation of Euclidean Algorithm in R 

<div class="column-left">
```{r eucalg2, echo = TRUE}

extGCDbigz <- function (a, b, showTrace = FALSE) 
{
  u <- c(as.bigz(1), as.bigz(0), as.bigz(a))
  v <- c(as.bigz(0), as.bigz(1), as.bigz(b))
  if(showTrace)
      cat(paste("v =", v[1], v[2], v[3], "\n"))
    while (v[3] != 0) {
      # Invariant: v[1]*a + v[2]*b = v[3]
      q <- u[3] %/% v[3]
      t <- u - v * q
      u <- v
      v <- t
      if(showTrace)
        cat(paste("v =", v[1], v[2], v[3], "\n"))
    }
    return(list(g = u[3], c = u[1], d = u[2]))
}
```
</div>

<div class="column-right">
```{r eucalgexec, echo = TRUE}
extGCDbigz(1180,482, TRUE)
```
</div>

## How fast is the extended Euclidean algorithm? 

```{r bigexample1, echo = TRUE}
extGCDbigz(672348327378426,24236368482, TRUE)
```

## How fast is the extended Euclidean algorithm? 

```{r bigexample2, echo = TRUE}
extGCDbigz(672348327378326,242363684823423, TRUE)
```

## Punchline: It's fast

The extended Euclidean algorithm runs in linear time on the number of digits (i.e., logarithmic time on the size of the numbers.)

## Finding inverses in $\mathbb{Z}_n$ 

The extended Euclidean algorithm gives us a way to find inverses in $\mathbb{Z}_n$. Suppose that $\gcd(a,n)=1$. Then we can use the extended Euclidean algorithm to find integers $c$ and $d$ such that $ca + dn = 1$. This gives us the multiplicative inverse of $a$. (How?)

>- $ca = 1-dn = 1 \mod n$
>- So $c = a^{-1}$ in $\mathbb{Z}_n$

---

```{r}
a <- as.bigz("908234890234894325723983248234892349")
n <- 256
gcd(a,n)
gcdex(a,n) # gmp version of extended Euclidean algorithm
(21 * a) %% n
```

## Inverses using `gmp`

There is also a `gmp` function that gives inverses in $\mathbb{Z}_n$ directly:

```{r}
inv.bigz(a,n) # gmp function for inverse in  Z_n
```

## What if $\gcd(a,n)\neq 1$?

Suppose $a$ and $n$ are *not* relatively prime. Then there is a common divisor $d > 1$ such that $a = kd$ and $n = ld$.

- Then $kld = kn = 0 \mod n$.
- But $kld = al$.
- So $al = 0 \mod n$, which makes $a$ a zero divisor in $\mathbb{Z}_n$.
- So $a$ is not invertible.

## Group Exercise

> Each function $s:\mathbb{Z}_{26}\longrightarrow \mathbb{Z}_{26}$ below defines an affine cipher. For each function, find the decryption function $s^{-1}$ or prove that no such function exists.
>
> 1. $x \stackrel{s}{\longmapsto} 11 x + 4$
> 
> 2. $x \stackrel{s}{\longmapsto} 13 x + 5$

Work in your group's voice channel with your group's Jamboard.

# Assignments

## Programming Assignment: due Sunday, 11:59 pm 

https://djhunter.github.io/cryptography/programming/01affine.html

## Written Assignment: due Monday, 11:59 pm 

https://djhunter.github.io/cryptography/assignments/02shiftaffine.html


