---
title: "Linear Feedback Shift Registers"
date: "September 17, 2020"
output: 
  revealjs::revealjs_presentation:
    fig_width: 14 
    fig_height: 7
    self_contained: true
    theme: night
    highlight: zenburn
    css: slidesdjh.css
    center: false
    transition: slide
    reveal_options:
      controls: true
      progress: false
      width: 1080
      height: 540
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(comment = NA)
options(width = 100)
library(gmp)
stringToMod26 <- function(x) {utf8ToInt(x)-utf8ToInt("a")}
mod26ToString <- function(x) {intToUtf8(x+utf8ToInt("a"))}
```

# $\mathbb{Z}_2$

## One-Time Pad

Plaintext: a vector $\mathbf{p} \in \mathbb{Z}_{26}^{N}$, where $N$ is the number of characters in the plaintext.

Key: a vector $\mathbf{k} \in \mathbb{Z}_{26}^{N}$, where $N$ is the number of characters in the plaintext.

Encryption function: $\mathbf{p} \longmapsto \mathbf{p} + \mathbf{k}$

Pros: Unbreakable

Cons: Big key

https://www.vice.com/en_us/article/ezvyyj/cubas-mysterious-numbers-station-is-still-on-the-air

```
!soundcloud https://soundcloud.com/wrench86/cuban-numbers-station-5-883-am
```

## [Ascii](http://www.asciitable.com/) and Binary {.smaller}

| Decimal | Hexadecimal | Binary | Character |
| :------ | :---------- | :----- | :-------- |
| 94 |	5E |	01011110 |	^ |
| 95 |	5F |	01011111 |	_ |
| 96 |	60 |	01100000 |	` |	
| 97 |	61 |	01100001 |	a |	 
| 98 |	62 |	01100010 |	b |	 
| 99 |	63 |	01100011 |	c |	 
| 100 |	64 |	01100100 |	d |	 
| 101 |	65 |	01100101 |	e |	 
| 102 |	66 |	01100110 |	f |	 
| 103 |	67 |	01100111 |	g |	 
| 104 |	68 |	01101000 |	h |	 
| 105 |	69 |	01101001 |	i |	 
| 106 |	6A |	01101010 |	j |	 
| 107 |	6B |	01101011 |	k |

## Binary and Hexadecimal in R

<div class="column-left">

```{r}
charToRaw("a")
charToRaw("abc")
rawToBits(charToRaw("a")) # note: backwards
```

</div>
  
<div class="column-right">

| Decimal | Hex | Binary | Char |
| :------ | :---------- | :----- | :------ |
| 94 |	5E |	01011110 |	^ |
| 95 |	5F |	01011111 |	_ |
| 96 |	60 |	01100000 |	` |	
| 97 |	61 |	01100001 |	a |	 
| 98 |	62 |	01100010 |	b |	 
| 99 |	63 |	01100011 |	c |	 
| 100 |	64 |	01100100 |	d |	 
| 101 |	65 |	01100101 |	e |	 
| 102 |	66 |	01100110 |	f |	 
| 103 |	67 |	01100111 |	g |	 
| 104 |	68 |	01101000 |	h |	 
| 105 |	69 |	01101001 |	i |	 
| 106 |	6A |	01101010 |	j |	 
| 107 |	6B |	01101011 |	k |

</div>

## Bitwise Logical Operators

<div class="column-left">

```{r}
charToRaw("a") & charToRaw("b") # bitwise and
charToRaw("a") | charToRaw("b") # bitwise or
xor(charToRaw("a"),charToRaw("b")) # bitwise xor
```

</div>

<div class="column-right">

| Decimal | Hex | Binary | Char |
| :------ | :---------- | :----- | :------ |
| 94 |	5E |	01011110 |	^ |
| 95 |	5F |	01011111 |	_ |
| 96 |	60 |	01100000 |	` |	
| 97 |	61 |	01100001 |	a |	 
| 98 |	62 |	01100010 |	b |	 
| 99 |	63 |	01100011 |	c |	 
| 100 |	64 |	01100100 |	d |	 
| 101 |	65 |	01100101 |	e |	 
| 102 |	66 |	01100110 |	f |	 
| 103 |	67 |	01100111 |	g |	 
| 104 |	68 |	01101000 |	h |	 
| 105 |	69 |	01101001 |	i |	 
| 106 |	6A |	01101010 |	j |	 
| 107 |	6B |	01101011 |	k |

</div>

## The ring $\mathbb{Z}_2$ {transition="slide-in fade-out"}

| Operation in $\mathbb{Z}_2$ | Operation on bits |
| :------ | :------ |
| Addition |  |
| Multiplication |  | 
| Subtraction |  | 

## The ring $\mathbb{Z}_2$ {transition="fade-in fade-out"}

| Operation in $\mathbb{Z}_2$ | Operation on bits |
| :------ | :------ |
| Addition | XOR |
| Multiplication |  | 
| Subtraction |  | 

## The ring $\mathbb{Z}_2$ {transition="fade-in fade-out"}

| Operation in $\mathbb{Z}_2$ | Operation on bits |
| :------ | :------ |
| Addition | XOR |
| Multiplication | AND | 
| Subtraction |  | 

## The ring $\mathbb{Z}_2$ {transition="fade-in slide-out"}

| Operation in $\mathbb{Z}_2$ | Operation on bits |
| :------ | :------ |
| Addition | XOR |
| Multiplication | AND | 
| Subtraction | XOR | 

Bonus feature: In $\mathbb{Z}_2$, $-1 = 1$. So you don't need negative signs!

## $\mathbb{Z}_2$ is a Field

- A field is a ring where every nonzero element has an inverse (and multiplication is commutative).
- Examples: $\mathbb{Q}$, $\mathbb{R}$, $\mathbb{Z}_p$, $p$ prime.
- Can add, subtract, multiply, and *divide*.
- Lots of good theorems.

## Punchline: Implement OTP in $\mathbb{Z}_2$

- Plaintext: a vector $\mathbf{p} \in \mathbb{Z}_{2}^{N}$, where $N$ is the number of characters in the plaintext.
- Key: a vector $\mathbf{k} \in \mathbb{Z}_{2}^{N}$, where $N$ is the number of characters in the plaintext.
- Encryption function: $\mathbf{p} \longmapsto \mathbf{p} + \mathbf{k}$

```{r}
oneTimePad <- function(plaintext, key) # needs fixing
{
  ptHex <- charToRaw(plaintext)
  keyHex <- charToRaw(key)
  ctHex <- xor(ptHex, keyHex[seq_along(ptHex)])
  return(rawToChar(ctHex))
}
```

## Null characters are a problem

```{r, eval=FALSE}
oneTimePad("hello", "mellow")
oneTimePad("hello", "harry")
```

One way to fix: 

- First, compute `ctHex` as in the original `oneTimePad`.
- Before converting `ctHex` to a string, change the leftmost bit of each byte from a `0` to a `1`. 
- Take this conversion into account when decrypting.

# Linear Feedback Shift Registers

## Fibonacci Numbers in $\mathbb{Z}_2$

The Fibonacci numbers are given by the following recurrence relation.
$$
F(n) = \left\{ \begin{array}{ll}
1 & \mbox{if $n=1$ or $n=2$} \\
F(n-1)+F(n-2) & \mbox{if $n>2$}
\end{array}\right.
$$

Interpreted as a function $F: \mathbb{N} \longrightarrow \mathbb{N}$, we get the following sequence of natural numbers:
$$
1, 1, 2, 3, 5, 8, 13, 21, 34, 55, 89, \ldots
$$

Interpreted as a function $F: \mathbb{N} \longrightarrow \mathbb{Z}_2$, we get the following sequence of binary digits:
$$
1, 1, 0, 1, 1, 0, 1, 1, 0, 1, 1, \ldots
$$

## Linear recurrence

In any ring, a **linear recurrence**  of length $m$ is a recursively-defined sequence of the form
$$
x_{n+m} = c_0x_n + c_1x_{n+1} + c_2x_{n+2} + \ldots + c_{m-1}x_{n+m-1}
$$
where $x_1, x_2, \ldots, x_n$ are given **initial values.** Note: We require $c_0 \neq 0$.

For example, the Fibonacci sequence is a linear congruence of length 2.
$$ 
x_{n+2} = x_n+x_{n+1}
$$
with $x_1 = 1 = x_2$.

## Linear recurrences can look "random" {.smaller}

For example, let $x_{n+31} = x_n + x_{n+3}$, and let the first 31 digits be 
$1 0 0 1 1 1 0 1 0 1 1 0 1 0 0 0 0 1 1 0 1 1 0 0 0 1 0 1 1 0 1$. The period of the resulting sequence is more than 2 billion.

```{r}
x <- numeric(1000)
x[1:31] <- c(1,0,0,1,1,1,0,1,0,1,1,0,1,0,0,0,0,1,1,0,1,1,0,0,0,1,0,1,1,0,1)
for(i in 1:969) {
  x[i+31] <- (x[i] + x[i+3]) %% 2 
}
x
```

## Period of linear recurrences

- The maximum period of a length $m$ linear recurrence in $\mathbb{Z}_2$ is $2^m-1$. (why?)
- [Theorem](http://www.ams.org/journals/mcom/1965-19-090/S0025-5718-1965-0184406-1/S0025-5718-1965-0184406-1.pdf): This maximum is achieved for recurrences 
$$x_{n+m} = c_0x_n + c_1x_{n+1} + c_2x_{n+2} + \ldots + c_{m-1}x_{n+m-1}$$
for which the polynomial
$$
p(x) = 1 + c_{m-1}x + c_{m-2}x^2 + \cdots + c_{1}x^{m-1} + x^m
$$
is [primitive](http://www.partow.net/programming/polynomials/primitive_polynomials_GF2.txt) over $\mathbb{Z}_2$. 

## Linear recurrences and OTP

A more practical OTP?

- Agree on a linear recurrence (i.e., A Linear Feedback Shift Register) with large period.
- Share an initial vector.
- Use the LFSR to generate a key for a one-time pad.
- ???
- Profit!

## How to hack a linear recurrence

Given a segment of bits from a linear recurrence, can we recover its formula?

Example: 

```{r, echo=FALSE}
x <- numeric(32)
x[1:5] <- c(1,0,0,1,1)
for(i in 1:27) {
  x[i+5] <- (x[i] +x[i+3]) %% 2 
}
sm2 <- matrix(c(x[1:2],x[2:3]), nrow=2)
sm3 <- matrix(c(x[1:3],x[2:4],x[3:5]), nrow=3)
cat(x)
```

| 1 | 0 | 0 | 1 | 1 | 0 | 1 | 0 | 0 | $\cdots$ |
| :--- | :--- | :--- | :--- | :--- | :--- | :--- | :--- | :--- | :--- |
| $x_0$ | $x_1$ | $x_2$ | $x_3$ | $x_4$ | $x_5$ | $x_6$ | $x_7$ | $x_8$ | $\cdots$ | 

1. Suppose $x_{n+2} = c_0x_n + c_1x_{n+1}$. Try to solve system of 2 equations. (me)
2. Suppose $x_{n+3} = c_0x_n + c_1x_{n+1} + c_2x_{n+2}$. Try to solve system of 3 equations. (You. Don't actually solve it, but try to write it as a matrix equation.)
3. etc.

## Another way to check invertibility

> Theorem. An $n \times n$ matrix over a field is invertible if and only if its determinant is nonzero.

(examples)

## Determinants in R

```{r}
sm2
det(sm2) %% 2
```

## Determinants in R

```{r}
sm3
det(sm3) %% 2
```

## Inverses
```{r}
sm3inv <- solve(sm3) %% 2
sm3inv
(sm3inv %*% sm3) %% 2
```

## Join your table group

```{r, echo=FALSE}
library(knitr)
roster <- c("Sam", "Wesley", "Andrew", "Winston", "Michael", "Logan", "Christian", "Caleb", "James", "Joe", "Emily", "Nathan")
set.seed(9172020) 
n <- length(roster)
ngps <- 3
maxingp <- ceiling(n/ngps)
# just make random groups
groups <- matrix(c(roster[sample(n)], 
                   rep("",(maxingp - (n %% maxingp)) %% maxingp)), 
                 ncol=maxingp, byrow=FALSE)
rownames(groups) <- paste0("Table #", 1:nrow(groups))
kable(t(groups))
```

Work on your group's [Jamboard](https://jamboard.google.com/d/1i8facJo6uJHeLJYADdMwPq2rjNI4wKLeEiILDWXybpw/edit?usp=sharing). You may want to upload a photo of your written work.

## Did it work?

**Group exercise:** Use the inverse matrix to solve for $c_0, c_1, c_2$, and check to see if it gives you the right recurrence relation.

1. Solve the matrix equation for $c_0$, $c_1$, $c_2$.
2. Compute the sequence defined by $x_{n+3} = c_0x_n + c_1x_{n+1} +c_2x_{n+2}$, starting with `100`. Do we get our given sequence?

## Binary Sequences and Shift Matrices

From a binary sequence, we can form an $n\times n$ matrix by taking the $i$th row to be the next $n$ terms of the sequence, starting at $i$. For the sequence $1, 1, 0, 1, 1, 0, 1, 1, 0, 1, 1, \ldots$, we get the following **shift matrices.**
$$
\begin{bmatrix}
1 & 1 \\
1 & 0
\end{bmatrix}, \quad
\begin{bmatrix}
1 & 1 & 0 \\
1 & 0 & 1 \\
0 & 1 & 1 
\end{bmatrix}, \quad
\begin{bmatrix}
1 & 1 & 0 & 1 \\
1 & 0 & 1 & 1 \\
0 & 1 & 1 & 0 \\
1 & 1 & 0 & 1
\end{bmatrix}, \quad
\begin{bmatrix}
1 & 1 & 0 & 1 & 1\\
1 & 0 & 1 & 1 & 0\\
0 & 1 & 1 & 0 & 1\\
1 & 1 & 0 & 1 & 1\\
1 & 0 & 1 & 1 & 0
\end{bmatrix}
$$

Which of these are invertible?

## Theorem: Linear recurrences are weak

> **Weakness Theorem.** Let $b_1, b_2, b_3, \ldots$ be a sequence of bits generated by a linear congruence. Suppose that the shortest linear congruence in $\mathbb{Z}_2$ that generates this sequence has the form $x_{n+m} = c_0x_n + c_1x_{n+1} + c_2x_{n+2} + \ldots + c_{m-1}x_{n+m-1}$. Let $S_N$ be the $N\times N$ shift matrix formed from the sequence. Then $S_m$ is invertible, and $S_N$ is not invertible for $N>m$. 

## Given bit string, find recurrence

1. Given bit string, form shift matrices $S_N$, and find the last $N$ that gives an invertible matrix.
2. Solve a system of equations to find the coefficients of the recurrence.

*Example:*

```{r, echo=FALSE}
x <- numeric(32)
x[1:5] <- c(1,0,0,1,1)
for(i in 1:27) {
  x[i+5] <- (x[i] +x[i+3]) %% 2 
}
cat(x)
```

## Shift matrices {.smaller}

```{r}
sm2 <- matrix(c(x[1:2],x[2:3]), nrow=2)
sm3 <- matrix(c(x[1:3],x[2:4],x[3:5]), nrow=3)
sm4 <- matrix(c(x[1:4],x[2:5],x[3:6],x[4:7]), nrow=4)
sm5 <- matrix(c(x[1:5],x[2:6],x[3:7],x[4:8],x[5:9]), nrow=5)
sm6 <- matrix(c(x[1:6],x[2:7],x[3:8],x[4:9],x[5:10],x[6:11]), nrow=6)
sm7 <- matrix(c(x[1:7],x[2:8],x[3:9],x[4:10],x[5:11],x[6:12],x[7:13]), nrow=7)
cat(det(sm2), det(sm3), det(sm4), det(sm5), det(sm6), det(sm7))
```

So it looks like our recurrence is length 5. Let $M$ be the 5th shift matrix. Then we need to solve 
$$
M \begin{bmatrix} c_0 \\ c_1 \\ c_2 \\ c_3 \\ c_4 \end{bmatrix} 
= \begin{bmatrix} 0 \\ 1 \\ 0 \\ 0 \\ 1 \end{bmatrix}
$$

## Multiply both sides by the inverse

$$
\begin{align}
M \begin{bmatrix} c_0 \\ c_1 \\ c_2 \\ c_3 \\ c_4 \end{bmatrix} 
&= \begin{bmatrix} 0 \\ 1 \\ 0 \\ 0 \\ 1 \end{bmatrix} \\
\begin{bmatrix} c_0 \\ c_1 \\ c_2 \\ c_3 \\ c_4 \end{bmatrix} &=
M^{-1}
\begin{bmatrix} 0 \\ 1 \\ 0 \\ 0 \\ 1 \end{bmatrix} 
= \begin{bmatrix} 1 \\ 0 \\ 0 \\ 1 \\ 0 \end{bmatrix}
\end{align}
$$

## R code to do this

```{r}
sm5inv <- round(solve(sm5) * 5) %% 2
(sm5inv %*% matrix(c(0,1,0,0,1), ncol = 1)) %% 2
```

Recurrence should be:
$$
x_{n+5} = x_n + x_{n+3}
$$

Check answer: Does the recurrence produce this sequence?
```{r echo=FALSE}
cat(x)
```

## Bonus: Compute the period

By the above theorem, the maximum period of a sequence produced by a length $m$ linear recurrence is $2^m - 1$. So the period of
```{r echo=FALSE}
cat(x)
```
must be $2^5-1 = 31$, because we observe no repeats in the first 31 digits.

Another way to see this is to look at the associated polynomial $p(x) = 1 + x^2 + x^5$. Since this polynomial is primitive over $\mathbb{Z}_2$, the period must be maximal.

Just for fun, here's a [list of primitive polynomials](http://www.partow.net/programming/polynomials/primitive_polynomials_GF2.txt) over $\mathbb{Z}_2$.



