---
title: "Attacks on RSA and Primality Testing"
date: "September 29, 2022"
output: 
  revealjs::revealjs_presentation:
    fig_width: 14 
    fig_height: 7
    self_contained: true
    theme: night
    highlight: zenburn
    css: slidesdjh.css
    center: false
    transition: slide
    reveal_options:
      controls: true
      progress: false
      width: 1080
      height: 540
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(comment = NA)
options(width = 100)
library(gmp)
library(openssl)
stringToBigz <- function(txt) {
  nraw <- charToRaw(txt)
  l <- length(nraw)
  return(sum(as.bigz(256)^(l-(1:l))*as.numeric(nraw)))
}
stringToBigz("Cubs!")

bigzToString <- function(n) {
  numbytes <- ceiling(log2.bigz(n)/8)
  nnumeric <- numeric(numbytes)
  for(i in 0:(numbytes-1)) {
    b <- as.numeric(mod.bigz(n, 256))
    n <- divq.bigz(n, 256)
    nnumeric[numbytes-i] <- b
  }
  return(rawToChar(as.raw(nnumeric)))
}
```

# Prime numbers

## Implementing RSA

How can we generate large prime numbers?

1. Choose a large, odd, random number.
2. Test to see if it is prime. If not, add 2 and repeat.

. . . 

The hard part is going to be determining if a large number is prime.

## Some prime number theorems

- There are infinitely many primes. (Euclid, 300 BC) Proof by contradiction.

. . .

- **The Prime Number Theorem.** (Hadamard, 1896) 
    - The $n$th prime is approximately (asymptotically) $n \ln n$. 
    - The number of primes less than $x$ is approximately (asymptotically) $\displaystyle{\frac{x}{\ln x}}$.
    - [Short proof](https://people.mpim-bonn.mpg.de/zagier/files/doi/10.2307/2975232/fulltext.pdf)
    
. . .

$$
\lim_{x \rightarrow \infty} \frac{\pi(x)}{\ln x} = 1
$$
    
## Testing primality

>- Exhaustive search for factors:
>    - Keep a big list of primes. 
>    - To determine if a given $n$ is prime, test all the prime numbers up to $\sqrt{n}$ to see if any divide $n$.
>- *Exercise:* If $n$ has 100 digits, how big will this list need to be (roughly)? 
>    - Fact needed: The number of primes less than $x$ is approximately $\displaystyle{\frac{x}{\ln x}}$.

## Recall Euler's Theorem

> **Euler's Theorem:** If $\gcd(a,n)=1$, then $a^{\phi(n)} \equiv 1 \pmod n$.

An immediate corollary is known as **Fermat's Little Theorem**. (FlT)

> If $p$ is prime, then $a^{p-1} \equiv 1 \pmod{p}$, for any $a < p$. 

## Contrapositive of FlT

```{r}
bigNumber <- as.bigz("2759906923170465870349284572513444495119639631
                     59159755581777302729039253474767543706477020042
                     97818745287627330276222679878256760119559870281
                     13146956853126194280414475892676777659939564742
                     44785168959704762122652461722078612423678394923
                     32839718788654764956662118459134315194058852873
                     11045581942871881388186309640860418454846129570
                     91247592793784631777384902156354032599108753489
                     93476673140050791456486240379352438607931317122
                     29805074199522231082822668601073716264370514378
                     45187017123708751530264674793463510019061497797
                     88166169654875208818278825075813161445176049456
                     40971682725620759622798463427777619513770932520
                     2107521")
powm(17, bigNumber-1, bigNumber)
```

What can you conclude?

## What can you conclude?

```{r}
powm(2, 562, 563)
powm(2, 560, 561)
```

## Fermat Primality Test

To test to see if $n$ is prime, 

1. Choose $a < n$.
2. Compute $x = a^{n-1} \bmod n$.
3. If $x\neq 1$, then $n$ is composite. If $x=1$ then $n$ is "probably" prime.

## Fermat (Weak) Pseudoprimes {.build}

A probable prime that isn't prime is called a "pseudoprime".

More precisely, if $a^{n-1} \equiv 1 \pmod{n}$, and $n$ is composite, then $n$ is a *(weak) pseudoprime for the base $a$.* 

For example, 561 is a pseudoprime for the base 2. Pseudoprimes make up a tiny fraction of any range of numbers (but not tiny enough).

## Square roots of 1?

How many $x$ satisfy $x^2 = 1$ in...

- $\mathbb{Z}_5$?
- $\mathbb{Z}_6$?
- $\mathbb{Z}_7$?
- $\mathbb{Z}_8$?

## Only two square roots of 1 in $\mathbb{Z}_p$

> **Lemma 1.** If there is some $x$ in $\mathbb{Z}_n$ such that $x^2 = 1$ but $x\neq \pm 1$, then $n$ is composite.

Sketch Proof: Let $d = \mbox{gcd}(x - 1, n)$. Show that $d \neq 1$.

## Improving the Fermat Primality Test {.build}

> **Lemma 1.** If there is some $x$ in $\mathbb{Z}_n$ such that $x^2 = 1$ but $x\neq \pm 1$, then $n$ is composite.

*Example.* $42^{204} \equiv 1 \pmod{205}$, so $205$ is a pseudoprime for the base $42$.

Note that $204 = 51 \cdot 2 \cdot 2$. Instead of raising to the 204th power all at once, we proceed bit by bit:

$$
\begin{aligned}
42^{51} &\equiv 83 \pmod{205} \\
83^{2} &\equiv 124 \pmod{205} \\
124^{2} &\equiv 1 \pmod{205}
\end{aligned}
$$

Apply lemma.

## Miller-Rabin Primality Test {.smaller}

To test to see if $n$ is prime, write $n-1 = 2^km$, where $m$ is odd. Choose $a<n-1$ randomly. The following computations are done in $\mathbb{Z}_n$. 

Let $b_0 = a^m$.

- If $b_0 = \pm 1$, then $n$ is probably prime. Stop.

Otherwise, let $b_1 = b_0^2$.

- If $b_1 = 1$, then $n$ is composite. Stop.
- If $b_1 = -1$, then $n$ is probably prime. Stop.

Otherwise, let $b_2 = b_1^2$.

- If $b_2 = 1$, then $n$ is composite. Stop.
- If $b_2 = -1$, then $n$ is probably prime. Stop.

Continue, if necessary, until reaching $b_{k-1}$. If $b_{k-1}\neq -1$, then $n$ is composite. Otherwise, $n$ is probably prime.

## Group Exercise

Step through Miller-Rabin for 561.

## Strong Pseudoprimes

If a composite $n$ is found to be probably prime by the Miller-Rabin test using the number $a$, it is called a **strong pseudoprime** for the base $a$. 

Strong pseudoprimes are very, very rare. So repeated applications of Miller-Rabin will determine primality with such a low probability of error as to be considered zero. According to the OpenSSL specification,

> `Both BN_is_prime_ex()` and `BN_is_prime_fasttest_ex()` perform a Miller-Rabin probabilistic primality test with `nchecks` iterations. If `nchecks == BN_prime_checks`, a number of iterations is used that yields a false positive rate of at most $2^{-80}$ for random input.

# More RSA considerations 

## Attacks on RSA

An RSA implementation is vulnerable if 

- $n$ can be easily factored. (factorization attack)
- $e$ or $d$ is too small. (low exponent attack)
- $m$ is too small. (short plaintext attack)
- Eve can observe how long decryption takes. (timing attack)

## Factorization Attacks

**Fermat's Factorization Method:** 

Suppose $n=pq$ and $p$ and $q$ are approximately the same size.

1. Calculate $n+i^2$ for $i=1,2,3,\ldots$ until you get a perfect square.

2. If $n+i^2 = k^2$, then $n=k^2-i^2= (k-i)(k+i)$.

*Prevention:* Check that $p$ and $q$ are far enough apart. ($10^{100}$)

## Exponent factorization attack

Step through the Miller-Rabin test for some choice of $a$. If at some point $b_i = 1$ for $i>0$, then $\gcd(b_{i-1}, n)$ is a factor of $n$.

Write $n-1 = 2^km$, where $m$ is odd. Let $b_0 = a^m$.

- If $b_0 = \pm 1$, then $n$ is probably prime. Stop.

Otherwise, let $b_1 = b_0^2$.

- If $b_1 = 1$, then $n$ is composite. Stop. **GCD is a factor.**
- If $b_1 = -1$, then $n$ is probably prime. Stop.

Otherwise, let $b_2 = b_1^2$.

- If $b_2 = 1$, then $n$ is composite. **GCD is a factor.**
- If $b_2 = -1$, then $n$ is probably prime. Stop. etc.

## Quadratic Sieve

> **Lemma 2.** If there are some $x$ and $y$ in $\mathbb{Z}_n$ such that $x^2=y^2$ but $x\neq \pm y$, then $n$ is composite, and $\gcd(x-y,n)$ is a factor.

Idea: Search for $x$ and $y$ by looking at products of squares of small primes.

Prevention: Make sure $p-1$ and $q-1$ don't have small prime factors.

## Low Exponent Attacks

If the exponent $e$ is chosen too small, and Eve can observe the same $m$ encrypted with several different keys $n_1, n_2, \ldots$, then Eve can solve a system of equations for $m$.
$$
\begin{aligned}
c_1 &= m^e \bmod n_1 \\
c_2 &= m^e \bmod n_2 \\
\vdots \\
\end{aligned}
$$
Only possible if $m^e$ is less than the product of the $n_i's$.

Prevention: Use "large" $e$. (65537)

## Short plaintext attack

If the plaintext is known to be a small number, Eve can use brute force to find it by encrypting all the small plaintexts. Even better, in $\mathbb{Z}_n$

- Make a list of $cx^{-e}$ for $x = 1, 2, \ldots$.
- Make a list of $y^e$ for $y = 1, 2, \ldots$.

If Eve finds a match, then $c = (xy)^e$, so $xy$ is the plaintext.

Prevention: padding.


## Padding

- Block ciphers work one block at a time.
- Some block cipher modes (e.g., ECB, CBC) also work one block at a time.
- If the plaintext is not an integral number of blocks, you need to pad it to make it so.
    - How do you decrypt? Make the padding say how many bytes you added. (e.g. `03 03 03`)
    - Decrypt, then throw away padding.
    
. . .

- Problem: if the plaintext *is* an integral number of blocks, and you don't pad it, decryption might think it has to throw away padding.
    - Solution: pad an integral number of blocks with a whole block of padding.
    - i.e., you always pad, so you always throw away padding when decrypting.

## Programming assignment

Returning [a bunch of things](../programming/05rsa.Rmd) in a list:

```{r, eval=FALSE}
makeRSAkey <- function() {
  p <- NA # TODO
  q <- NA # TODO
  e <- as.bigz("65537")
  d <- NA # TODO
  return(list(n=p*q, p=p, q=q, e=e, d=d))
}

```


## Written assignment

- [Written Assignment](../assignments/10primes.html)

