---
title: "Discrete Logarithms, Diffie-Hellman, and ElGamal"
date: "October 14, 2022"
output: 
  revealjs::revealjs_presentation:
    fig_width: 14 
    fig_height: 7
    self_contained: true
    theme: night
    highlight: zenburn
    css: slidesdjh.css
    center: false
    transition: slide
    reveal_options:
      controls: true
      progress: false
      width: 1080
      height: 540
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(comment = NA)
options(width = 100)
library(gmp)
library(openssl)
```

# Groups and Generators

## The group of units of $\mathbb{Z}_n$

The set of all invertible elements of the ring $\mathbb{Z}_n$ is denoted $U(n)$, and is called **the group of units** of $\mathbb{Z}_n$. 

> Exercise: Find all the elements of $U(8)$, and make a multiplication table. Is U(8) a ring? Why or why not?

## Groups, cyclic groups, generators

A set with (an associative) multiplication and multiplicative inverses is called a **group**. For example, $U(n)$ is a group.

If all the elements of a group $G$ are powers of a single element $\alpha$ of the group, then $G$ is called a **cyclic group**, and $\alpha$ is called a **generator** of the group $G$.

> Exercise: Which of the following are cyclic groups? 
>
> - $U(8)$
> - $U(10)$
> - $U(7)$

## The Primitive Root Theorem

**Theorem.** The group $U(n)$ is cyclic if and only if $n$ is $1$, $2$, $4$, $p^k$, or $2p^k$, where $p$ is an odd prime and $k\geq 1$.

For a proof, see [Guichard], *Mathematics Magazine* 72:2, April 1999, pp. 139--142.

When $U(n)$ is cyclic, a generator of $U(n)$ is called a **primitive root** modulo $n$.

## Order of an element in a group

**Definition.** Let $G$ be a finite group, and let $g\in G$. The **order** of the element $g$ is the smallest number $e$ such that $g^e = 1$.

**Theorem.** The order of an element in a finite group divides the size of the group.

Proof: Math 110

(Example)

## Finding Primitive Roots modulo $p$ {.build .smaller}

- Brute Force
- Using the factorization of $p-1$ (See also Exercise 21, p. 107)

Example: Find a generator of $U(43)$ (i.e., a primitive root modulo 43)

## Brute Force

Is 2 primitive in $U(43)$?

```{r}
powm(rep(2,42), 1:42, 43) # Brute force
```

Is 3 primitive in $U(43)$?

```{r}
powm(rep(3,42), 1:42, 43) # Brute force
```

## Shortcut to test primitivity

Suppose $p$ is prime and $p-1 = q_1^{i_1}q_2^{i_2}\cdots q_n^{i_n}$ is a prime factorization.

>- If $g^r = 1$ for some $r<p-1$, then $g$ is not primitive. If no such $r$ exists, then $g$ is primitive.
>- If $r<p-1$ divides $p-1$, then it has to divide $(p-1)/q_k$ for some $k$.
>- If $g^r = 1$ for some $r<p-1$, then $r$ must divide $p-1$, so $r$ must divide $(p-1)/q_k$ for some $k$.
>- If $g^r = 1$ and $r$ divides $(p-1)/q_k$, then $g^{(p-1)/q_k} = 1$.
>- So we only need to check powers of the form $g^{(p-1)/q_k}$ to determine whether $g$ is primitive.

## Use the factors of $p-1$

Is 2 primitive?

```{r}
powm(c(2,2,2), c(21,6,14), 43) # Using factorization 42 = 2*3*7
```

Is 3 primitive?

```{r}
powm(c(3,3,3), c(21,6,14), 43) # Using factorization 42 = 2*3*7
```

# Discrete Logarithms

## Diffie-Hellman Key Exchange

Alice and Bob wish to create a shared secret $K$ (e.g., to make an AES key) using unencrypted communication.

1. Alice and Bob agree on a choice of a large prime $p$ and a generator $\alpha \in U(p)$.
2. Alice chooses a secret random $x \in \{1,2,\ldots, p-2\}$. Bob chooses a secret random $y \in \{1,2,\ldots,p-2\}$.
3. Alice sends $\alpha^x$ to Bob. Bob sends $\alpha^y$ to Alice. (All computations are done in $U(p)$.)
4. Alice calculates $K = (\alpha^y)^x$. Bob calculates $K = (\alpha^x)^y$

In practice, an AES (or other) key is obtained by taking a "hash" of $K$. (More details in Chapter 8.)

## Discrete Logarithms

In the Diffie-Hellman key exchange, Eve can observe $\alpha^x$ and $\alpha^y$, but she can't obtain the secret $K$ unless she knows $x$ or $y$. So the security of this algorithm relies on the fact that, in $U(p)$, the equation
$$
\beta = \alpha^x
$$
is difficult to solve for $x$.

> **Definition.** Let $\alpha,\beta \in U(p)$, with $p$ prime. The least nonnegative integer $x$ that satisfies the equation $\beta = \alpha^x$ is called the **discrete logarithm** of $\beta$ with respect to $\alpha$, denoted $L_\alpha(\beta)$.

Example: In $U(11)$, compute $L_3(4)$.

## Attacks on Diffie-Hellman: Computing $L_\alpha(\beta)$

- Brute force 
- Pohlig-Hellman Algorithm
- Baby Step, Giant Step

## Discrete Logarithms modulo 2

Suppose $\alpha$ is a primitive root modulo $p$ and consider solving $\beta = \alpha^x$ in $U(p)$. (i.e., compute $L_\alpha(\beta)$ modulo $p$.)

The size of $U(p)$ is $p-1$. Write $p-1 = 2c$.

**Exercise 1.** All calculations below are in $U(p)$.

1. What is $\alpha^{2c}$? Why?
2. What is $\alpha^c$? Why?
3. Write $\beta^c$ in terms of $x$.

If $x$ is even, what must be true? What if $x$ is odd?

## Solving simultaneous congruences

**Exercise 2.** Find a value of $x$ that satisfies both of the following congruences.
$$
\begin{aligned}
x &\equiv 3 \pmod{5} \\
x &\equiv 2 \pmod{7}
\end{aligned}
$$

**Exercise 3.** Find a value of $x$ that satisfies both of the following congruences.
$$
\begin{aligned}
x &\equiv 3 \pmod{4} \\
x &\equiv 2 \pmod{6}
\end{aligned}
$$

## The Chinese Remainder Theorem

**Theorem.** Let $m_1, m_2, \ldots m_k$ be integers that are *pairwise relatively prime*, and let $M = m_1m_2\cdots m_k$. Given integers $a_1, a_2, \ldots, a_k$, there is a solution $x$ to the following simultaneous congruences. Furthermore, this solution is unique modulo $M$.
$$
\begin{aligned}
x &\equiv a_1 \pmod{m_1} \\
x &\equiv a_2 \pmod{m_2} \\
&\,\,\vdots \\
x &\equiv a_k \pmod{m_k} \\
\end{aligned}
$$

Constructive proof: Let $x = a_1y_1z_1 + a_2y_2z_2 + \ldots + a_ky_kz_k$, where $z_i = M/m_i$ and $y_i = z_i^{-1} \bmod m_i$.

## CRT: Example

**Exercise 4.** Find a value of $x$ that satisfies all of the following congruences.
$$
\begin{aligned}
x &\equiv 2 \pmod{5} \\
x &\equiv 1 \pmod{6} \\
x &\equiv 3 \pmod{7}
\end{aligned}
$$

## Applying the CRT

1. Solve your problem for a collection of small moduli $m_i$.
2. Use the CRT find the solution for the modulus $M = m_1m_2\cdots m_k$.

## Pohlig-Hellman Algorithm (overview)

Suppose $p-1 = m_1m_2\cdots m_k$, where each $m_i$ is a power of a small prime.

1. Compute $L_\alpha(\beta)$ modulo $m_i$, for each $i$.
2. Apply the Chinese Remainder Theorem to compute $L_\alpha(\beta)$.

Facts used:

- $\beta^{(p-1)/2}$ tells us $L_\alpha(\beta)$ mod 2 immediately (Exercise 1).
- $\alpha^{m_1} \equiv \alpha^{m_2} \pmod{p}$ iff $m_1\equiv m_2 \pmod{p-1}$. 

## Pohlig-Hellman Attack: Toy Example

Compute $L_7(12)$ in $U(41)$. (i.e., Solve $7^x = 12$.)

>- $p-1 = 41 - 1 = 40 = 2^3 \cdot 5$
>- Since $12^{20} \equiv -1 \pmod{41}$, we know $x = L_7(12) \equiv 1 \pmod{2}$.
>    - In fact, *$L_7(12) \equiv 5 \pmod{8}$*. (Details omitted.)
>- Observe: $12^{40/5} = 12^8 \equiv 18 \pmod{41}$, and $7^8 \equiv 37 \pmod{41}$.
>- Try powers of $37$ and find that $37^3 \equiv 18 \pmod{41}$.
>- So $(7^8)^3 \equiv 12^8 = (7^x)^8 \pmod{41}$. So $8\cdot 3 \equiv x \cdot 8 \pmod{40}$.
>- It follows that $3 \equiv x \pmod{5}$.
>- Therefore *$L_7(12) \equiv 3 \pmod{5}$*. 
>- Now use CRT: $L_7(12) = (5)(5)(5) + (3)(2)(8) \pmod{40} = 13$.

## Avoid the Pohlig-Hellman Attack?

How to avoid this attack: Make sure $p-1$ has at least one large prime factor.

```{r}
factorize(as.bigz("234804645645701") - 1)
```

## Baby Step, Giant Step

(Similar to meet-in-the-middle attack.)

To solve $\alpha^x = \beta$ in $U(p)$: Choose large $N > \sqrt{p-1}$

1. Compute $\alpha^j$ for $j = 1, 2, \ldots, N$. (Baby step.)
2. Compute $\beta\alpha^{-Nk}$ for $k = 1, 2, \ldots, N$. (Giant step.)

If a match is found in these two lists, $x=j+Nk$ is a solution. (Note that $k$ and $j$ are the "digits" in the base $N$ representation of $x$.)

How to avoid this attack: Use large primes.

## In practice: Use "good" Diffie-Hellman groups.

[Secure Diffie-Hellman Groups](https://datatracker.ietf.org/doc/rfc3526/?include_text=1)

## ElGamal public key cryptosystem

1. Bob chooses a large prime $p$ and a generator $\alpha\in U(p)$, as well as a secret exponent $a$. Bob computes $\beta = \alpha^a$ in $U(p)$, and makes the triple $(p,\alpha, \beta)$ public.
2. Alice chooses a secret random $k$ and computes $r = \alpha^k$ in $U(p)$.
3. To send the plaintext message $m$, Alice computes $t = \beta^km$ in $U(p)$ and sends the pair $(r,t)$ to Bob.

Decryption: In $U(p)$, Bob uses the secret $a$ to compute $tr^{-a}$. This works because
$$
tr^{-a} = \beta^km(\alpha^k)^{-a} = (\alpha^a)^km\alpha^{-ak} = m
$$




