---
title: "Hash Functions"
date: "October 15, 2020"
output: 
  revealjs::revealjs_presentation:
    fig_width: 14 
    fig_height: 7
    self_contained: true
    theme: night
    highlight: zenburn
    css: slidesdjh.css
    center: false
    transition: slide
    reveal_options:
      controls: true
      progress: false
      width: 1080
      height: 540
---

# Last Time: Primitive Roots

## Recall: Primitive Roots

- When $U(n)$ is cyclic, a generator of $U(n)$ is called a **primitive root** modulo $n$.
- In $U(p)$ where $p$ is prime, for any $a$, $a^{p-1} = 1$ (Fermat's Little Theorem).
- However, the powers $a, a^2, a^3, \ldots a^{p-1}$ are not always *distinct* elements of $U(n)$. But they are distinct exactly when $a$ is primitive.

*Programming Problem:* Automate your procedure in Problem #1 from last night: `nextPrimRoot` function.

## Recall: Discrete Logarithms are hard to compute

Suppose $\alpha$ is a primitive root modulo large prime $p$.

- Given $\alpha$ and $x$, it is easy to compute $\alpha^x$.
- Given $\alpha$ and $\alpha^x$, it is nearly impossible to compute $x$.

*Programming Problems:* Implement two discrete logarithm finders.

1. Brute force.
2. Baby-step, Giant-step

## Baby Step, Giant Step

To solve $\alpha^x = \beta$ in $U(p)$: Choose large $N > \sqrt{p-1}$

1. Compute $\alpha^j$ for $j = 1, 2, \ldots, N$. (Baby step.)
2. Compute $\beta\alpha^{-Nk}$ for $k = 1, 2, \ldots, N$. (Giant step.)

If a match is found in these two lists, $x=j+Nk$ is a solution. (Note that $k$ and $j$ are the "digits" in the base $N$ representation of $x$.)

**Advice:** Check out R's built-in `match` function. It doesn't work on `bigz`'s, but it works on strings, so you can `as.character()` your `bigz`'s to strings before you try using `match()`

## Using `match`

```{r}
w1 <- c("I", "thought", "you", "said", "your", "dog", "did", "not", "bite")
w2 <- c("That", "is", "not", "my", "dog")
m <- match(w1, w2)
m
which(!is.na(m))
commonWords <- w2[m]
commonWords
commonWords[!is.na(commonWords)]
```

## Hash functions

Given a set of data (e.g., an employee's name, address, birthday etc.) we would like to assign a "unique" identifier to the data set in a fast, algorithmic way. This identifier is called a **hash** or a **message digest** of the given set of data.

Applications:

>- Organizing data
>- Checking for data integrity
>- Finding duplicate records
>- Password checking
>- Digital signatures
>- Pseudorandom bit string generation
>- Generating a key (e.g. AES) from a passphrase

## Hash Functions

A **hash function** $h:M \longrightarrow D$ is a function that inputs a message of arbitrary length and outputs a **message digest** of fixed length. 

## Example: Checking data integrity

```{r}
library(openssl)
bg1 <- "Take me out to the ball game, Take me out with the crowd; Buy me some peanuts and 
        Cracker Jack, I don't care if I never get back. Let me root, root, root for the 
        home team, If they don't win, it's a shame. For it's one, two, three strikes, 
        you're out, At the old ball game."
bg2 <- "Take me out to the ball game, Take me out to the crowds;  Buy me some peanuts and 
        Cracker Jack, I don't care if I never get back. Let me root, root, root for the 
        home team, If they don't win, it's a shame. For it's one, two, three strikes, 
        you're out, At the old ball game."
md5(c(bg1,bg2))
```

## Example: Password verification

A system stores hashed passwords:

```{r}
passwords <- c("qwerty", "mypass", "admin", "Runner4567")
sha256(passwords) 
```

When a user enters a password $m$, the hash $h(m)$ is compared to the stored values.

## Properties of Cryptographic Hash Functions 

Cryptographic hash functions should satisfy the following properties:

>- **Efficiency.** Computing $h(m)$ is fast.
>- **One-way.** Given a digest $d$, it is practically impossible to find a message $m$ such that $h(m)=d$.
>- **Collision-free.** It is practically impossible to find messages $m_1$ and $m_2$ such that $h(m_1)=h(m_2)$.
>- **Diffusion.** Changing one bit of $m$ completely changes $h(m)$.

## Examples

Discuss. For each possible hash function $h$, which properties (efficiency, one-way, collision-free, diffusion) does $h$ have?

> 1. $h(m) = m \bmod p$, where $m > p$.
> 2. $h(m) = \alpha^m \bmod p$, where $\alpha$ is a primitive root modulo $p$.
> 3. $h(m) =$ the first block of the AES ciphertext of $m$ using a public key.

## Common Hash Functions

| Function | Year | Digest size | Secure? |
|----------|------|-------------|---------|
| MD5 | 1991 | 128 bit | no (2004-2005)|
| SHA-1 | 1993 | 160 bit | no (2005) |
| SHA-2 | 2001 | 224, 256, 384, or 512 bit | yes | 
| SHA-3 | 2015 | 224, 256, 384, or 512 bit | yes |

<br>

For example, openssl defaults to `sha256`, which is SHA-2. [Check out the pseudocode.](https://en.wikipedia.org/wiki/SHA-2#Pseudocode)

## Dictionary attacks

Suppose a set $P$ of passwords are stored as hashes: $\{h(p) \mid p \in P\}$.

An attacker can create a list $L = \{h(d) \mid d \in \mathcal{D}\}$, where $\mathcal{D}$ is a **dictionary** list of common passwords.

A match between $L$ and $P$ will reveal a password.

```{r}
passwords <- c("qwerty", "mypass", "admin", "Runner4567")
sha256(passwords) 
```

## Salt

Password systems sometimes incorporate a random string, called **salt**, into a password before it is hashed.

```
nyoung:$6$WQ0w1hcR$7wO.W0nyjzcakz42KgoPhfgdP8NWOZXSSTZzxEhPBTCiXAKcJ1kHv4VgTMKyx
U6Y7XNXaBmkPtlsLp1q7VPVC1:18212:0:99999:7:::
rfrink:$6$T3ortZMG$HOC3KWnrWHGvjujYnH42ZrfhdDVBbbfXLYFFG80rG9kvnJynEY8O/V37l.1T/
kcitXgkBFobngRxFYlhmXEL5.:18166:0:99999:7:::
roroku:$6$YTcauXgg$CFgWgR/oXj3NhFV56oqosnQAxrBto5Ho640uzTmcAR0OnRD.qkY4O6zbjVP
NqIoguUuLCK4Y8WW.PDmPWPK0:18227:0:99999:7:::
samundson:$6$CllynVdl$9N76VLgzfR1XsrqwSscksfgMOWs59oSW67ai4g3oHu29lnXe2pwk4g9Wvmp
A6XMHH8W0Sq4togY1I6UQKgeBMV0:18166:0:99999:7:::
```

*Question:* If $B$ is the number of bits of salt, how does salting affect the time needed for a dictionary attack?

## Pepper

- Similar to salt 
- Pepper is random and secret, while salt is just random.
- Store pepper somewhere else; salt gets stored with the hashed passwords.

## Random Oracle

An ideal hash function $h: M \longrightarrow D$ is a **random oracle**.

- The oracle answers the question $m\in M$ with a randomly selected $h(m) \in D$ (selected with replacement).
- If the same question $m$ is asked a second time, the same answer $h(m)$ is given.

To calculate theoretical probabilities, we often assume we have a random oracle hash function.

## The Birthday Paradox 

> What is the probability that two people in a group have the same birthday?

- Write down your birthday (day/month) and the birthdays of two family members.
- Did we obtain a collision?

```{r, echo=FALSE, eval=FALSE}
1-prod(1-(1:33)/365)
```

## The Birthday Theorem

**Birthday Theorem.**  Suppose $h:M \longrightarrow D$ is a hash function, where there are $N$ possible hashes. Assuming that $h$ is a random oracle, the probability of a collision when hashing $r$ messages is
$$
1-\left(\frac{N-1}{N}\right)\left(\frac{N-2}{N}\right)\cdots\left(\frac{N-r+1}{N}\right)
$$
This probability is approximately $1-e^{-r^2/2N}$.

```{r}
1-prod(1-(1:33)/365)
1-exp(-33^2/(2*365))
```

## Birthday Attacks 

Choose a value for $r$, and look for collisions.

How big must $r$ be to have a "high" probability $\gamma$ of finding a collision? Solve (asymptotically):

$$
1-e^{-r^2/(2N)} = \gamma
$$

$r \approx \sqrt{N}$, asymptotically. (Can solve exactly, given $\gamma$.)

"Number of bits of security" $\approx \frac{1}{2} \times \text{digest size}$

## Programming a birthday attack

*Programming problem:* Write a birthday attack on a semi-weak hash function `miniSHA`.

- Try hashing a bunch of strings, and find a collision (i.e., a duplicated hash for two different strings).
- **Advice:** Check out R's built-in `duplicated` function. Read the help menu to see how it works.
- Your hashes will be strings, so `duplicated` will work just fine.

## How `duplicated` works

The function `duplicated` "determines which elements of a vector or data frame are duplicates of elements with smaller subscripts, and returns a logical vector indicating which elements (rows) are duplicates."

```{r}
piDigits <- c(3, 1, 4, 1, 5, 9, 2, 6, 5, 3, 5)
dups <- duplicated(piDigits)
dups
which(dups)
piDigits[which(dups)]
```

## Application: Using hash functions to encrypt

Idea: create a random bitstring using hash functions, a key, and some sort of feedback. Encrypt by XORing (as with certain block cipher modes).

## Application: Merkle Tree data structure

Problem: verify the integrity of $n$ independent blocks $B_1, B_2, \ldots, B_n$. (e.g., peer-to-peer data)

- $h(B_1), h(B_2), \ldots, h(B_n)$ form the leaves of the tree.
- Each parent node is formed by concatenating the children and hashing.
- Only the top hash needs to be checked to verify integrity of all blocks.

## Application: Block chaining

Examples: Git, Bitcoin

- Each new block contains a hash of the previous block.
- Difficult to go back and change an old block, because the hashes won't check.

## Application: Bitcoin Proof of Work

Problem: A new block needs to be added to the Bitcoin blockchain. Who decides?

- Miners race to find a *nonce*, such that, when appended to the new block, the hash begins with a certain number of zeroes. 
- Anybody can check that the nonce works, to verify the integrity of the new block.
- The first one to find a nonce that works adds the new block to the blockchain and gets rewarded. Everybody is going to accept the new blockchain because it is the longest.

Reference: https://bitcoin.org/bitcoin.pdf




