---
title: "Elliptic Curves Mod p"
date: "October 27, 2022"
output: 
  revealjs::revealjs_presentation:
    fig_width: 14 
    fig_height: 7
    self_contained: true
    theme: night
    highlight: zenburn
    css: slidesdjh.css
    center: false
    transition: slide
    reveal_options:
      controls: true
      progress: false
      width: 1080
      height: 540
---

# Elliptic Curves modulo p

## Addition Law for Real Elliptic Curves 

If $P$ is any point on an elliptic curve, then $P + \infty = P = \infty + P$. Let $P_1 = (x_1,y_1)$ and $P_2 = (x_2,y_2)$ be points on the elliptic curve $y^2 = x^3 + bx +c$. 

If $P_1 = P_2$ and $y_1 = 0$, or if $P_1 \neq P_2$ and $x_1 = x_2$, then $P_1+P_2 = \infty$.

Otherwise, $P_1 + P_2 = P_3 =  (x_3,y_3)$, where
$$
\begin{aligned}
x_3 &= m^2 -x_1 -x_2 \\
y_3 &= m(x_1-x_3)-y_1
\end{aligned}
$$
and
$$
m = \left\{\begin{array}{cl}
(y_2-y_1)/(x_2-x_1) & \mbox{if } P_1 \neq P_2 \\
(3x_1^2+b)/(2y_1) & \mbox{if } P_1 = P_2 
\end{array}
\right.
$$
where all computations are performed in $\mathbb{R}$.

## Addition Law for Mod $p$ Elliptic Curves 

If $P$ is any point on an elliptic curve, then $P + \infty = P = \infty + P$. Let $P_1 = (x_1,y_1)$ and $P_2 = (x_2,y_2)$ be points on the elliptic curve $y^2 = x^3 + bx +c$. 

If $P_1 = P_2$ and $y_1 = 0$, or if $P_1 \neq P_2$ and $x_1 = x_2$, then $P_1+P_2 = \infty$.

Otherwise, $P_1 + P_2 = P_3 =  (x_3,y_3)$, where
$$
\begin{aligned}
x_3 &= m^2 -x_1 -x_2 \\
y_3 &= m(x_1-x_3)-y_1
\end{aligned}
$$
and
$$
m = \left\{\begin{array}{cl}
(y_2-y_1)(x_2-x_1)^{-1} & \mbox{if } P_1 \neq P_2 \\
(3x_1^2+b)(2y_1)^{-1} & \mbox{if } P_1 = P_2 
\end{array}
\right.
$$
where all computations are performed in $\mathbb{Z}_p$.


## Example: real vs. mod $p$

$y^2 = x^3 -2x + 10$, $P=(1,3)$

```{r, eval=FALSE}
> sapply(1:8, function(x){ecPowReal(-2,10,c(1,3),x)}) # real
     [,1]      [,2]      [,3]      [,4]      [,5]       [,6]       [,7]     [,8]
[1,]    1 -1.972222  4.402219  7.670201 -1.293236  0.3031689   74.56223 1.730672
[2,]    3 -2.504630 -9.300994 21.116645  3.228558 -3.0694506 -643.73252 3.423800

> sapply(1:8, function(x){as.character(ecPowModp(-2,10,8363,c(1,3),x))}) # mod 8363
     [,1] [,2]   [,3]   [,4]   [,5]   [,6]   [,7]   [,8]  
[1,] "1"  "3018" "3103" "3650" "1147" "7141" "1658" "1290"
[2,] "3"  "888"  "1130" "3640" "6505" "2966" "5256" "6565"
```


## Big idea

Mod $p$ elliptic curves give us groups to replace $U(p)$ groups.

- $U(p)$ uses *multiplicative* notation.
    - "Powers" in $U(p)$ are elements of the form $\alpha^n = \underbrace{\alpha \cdot  \alpha \cdot \cdots \cdot \alpha}_{n \; \alpha \mbox{'s}}$
    - The discrete logarithm problem was hard in $U(p)$.
- Mod $p$ elliptic curves use *additive* notation.
    - "Powers" in mod $p$ elliptic curves are elements of the form $nP = \underbrace{P + P + \cdots + P}_{n \; P \mbox{'s}}$
    - The discrete "logarithm" problem is *harder* in mod $p$ elliptic curve groups.

# Factorization revisited

## Factorization using gcd

- Suppose a number $n$ is composite (e.g., $n = pq$).
- If you can find some number $d$ such that $\gcd(d, n) \neq 1$, you can factor $n$.
- The art lies in choosing clever guesses for $d$.

## Pollard's $p-1$ method

Problem: Factor an integer $n$.

1. Choose a base $a$ between $2$ and $n-2$.

2. Choose a large $B$, and compute $r = a^{B!} \bmod n$.

3. If $\gcd(r-1,n) \neq 1$, then you have found a nontrivial factor of $n$.

Why does this method work? (Fermat)

## Pollard's method: Example

$n = 540143, a = 2, B = 7$

```{r, message=FALSE}
library(gmp)
r <- powm(2, factorialZ(7), 540143)
gcd(r-1, 540143)
```
```{r}
540143/421
```

## Pollard's $p-1$ method as a group traversal

Problem: Factor an integer $n$.

1. Choose a base $a$ between $2$ and $n-2$.

2. Choose a bound $B$, and for $k$ in $\{1!, 2!, 3!, \ldots B!\}$ attempt the following:
    a. Compute $r = a^k \bmod n$.
    b. If $r-1 \in U(n)$, try the next $k$ or another $a$
    c. If $r-1 \not\in U(n)$, then $\gcd(r-1,n)$ is a nontrivial factor of $n$.


## Pollard's method: Example

$n = 540143$

Choose $a = 2$ and $B = 8$.

$$
\begin{aligned}
(2^1 \bmod 540143) - 1 &= 1 \in U(540143) \\
(2^2 \bmod 540143) - 1 &= 3 \in U(540143) \\
(2^6 \bmod 540143) - 1 &= 63 \in U(540143) \\
(2^{24} \bmod 540143) - 1 &= 32782 \in U(540143) \\
(2^{120} \bmod 540143) - 1 &= 54804 \in U(540143) \\
(2^{720} \bmod 540143) - 1 &= 518076 \in U(540143) \\
(2^{5040} \bmod 540143) - 1 &= 167137 \not\in U(540143) 
\end{aligned}
$$

$\gcd(167137, 540143) = 421$, and $540143/421 = 1283$.

## Pollard's method in R 

```{r}
powers <- factorialZ(1:10)
gcd(powm(rep(2,10), powers, 540143)-1, 540143)
```

It worked because $421-1$ has no large prime factors:

```{r}
factorize(420)
```

## Large prime factors make it harder

```{r}
powers <- factorialZ(1:10)
gcd(powm(rep(2,10), powers, 455839)-1, 455839)
```

Here $455839 = 599 \cdot 761$, but $598 = 2 \cdot 3 \cdot 23$ and $760 = 2^3\cdot 5 \cdot 19$.

*Question:* How big a factorial would we need to use?

## Factorization: Lenstra's method

Idea: Replace $U(n)$ with mod $p$ elliptic curve groups

Problem: Factor an integer $n$.

1. Choose an elliptic curve $E$ and a point $P$ on the curve. 

2. Choose a bound $B$, and for $k$ in $\{1!, 2!, 3!, \ldots B!\}$ attempt the following:
    a. Attempt to compute $kP$ on $E$.
    b. If the computation succeeds for all $k$, try another $E$.
    c. If the computation fails because one of the denominators $d$ in the Addition Rule is zero, then $\gcd(d,n)$ is a nontrivial factor of $n$.

## Lenstra's method: Example

Factor $n = 455839$. Choose $E$ to be $y^2 = x^3+5x-5$ modulo $n$. Let $P = (1,1)$. 

$$
\begin{aligned}
2P &= (14,455786) \\
6P &= (179685,427131) \\
24P &= (67658,34840) \\
120P &= (129563, 249834) \\
720P &= (422839, 402728) \\
5040P &= (70028, 403526) \\
40320P & \mbox{ does not exist.}
\end{aligned}
$$

## Lenstra's method in R


```{r, eval=FALSE}
> ecPowModp(5, -5, 455839, c(1,1), 720)
Big Integer ('bigz') object of length 2:
[1] 422839 402728

> ecPowModp(5, -5, 455839, c(1,1), 5040)
Big Integer ('bigz') object of length 2:
[1] 70028  403526

> ecPowModp(5, -5, 455839, c(1,1), 40320)
Warning in ecAddModp(b, c, modulus, pow1, pow1): 315074 is not invertible mod 455839
Warning in inv.bigz((2 * y1), modulus): inv(x,m) returning NA as x has no inverse modulo m

Big Integer ('bigz') object of length 2:
[1] NA NA
```

Therefore $\gcd(315074,455839)=599$ is a nontrivial factor of $n$.

## Why did it work?

"Inside" an elliptic curve with modulus $pq$, you have two elliptic curves: one modulo $p$, and another modulo $q$. Note: $455839 = 599 \cdot 761$

```{r, eval=FALSE}
> ecPowModp(5, -5, 761, c(1,1), 40320)
Big Integer ('bigz') object of length 2:
[1] 464 650

> ecPowModp(5, -5, 599, c(1,1), 40320)
[1] Inf Inf
```

## Efficient Exponentiation

```{r pow, echo=TRUE}
rPower <- function(x, n)
{
  if(n==0) return(1)
  x*rPower(x, n-1)
}

qPower <- function(x, n)
{
  if(n==0) return(1)
  if(n %% 2 == 0)
    return((qPower(x, n %/% 2))^2)
  else
    return(x*(qPower(x, n %/% 2))^2)
}
```

>- **Programming assignment:** Implement fast powers for mod $p$ elliptic curves.

# Tools for using elliptic curves in cryptography

## Finding points on elliptic curves

*Group Exercise:* In $\mathbb{Z}_7$:

1. Find all the perfect squares.
2. Find all the elements of the form $x^3 + 2$.
3. Find all the points on the elliptic curve $y^2 = x^3 + 2$.

## Square roots mod $p$

How many perfect squares are there in $U(p)$?

## Theorem on square roots mod $p$ (p. 86)

**Theorem.** Let $p$ be prime. In $U(p)$, $y$ has a square root if and only if $y^{(p-1)/2} \equiv 1 \pmod{p}$.

In this case, if $p \equiv 3 \pmod{4}$, then
$$
\pm y^{(p+1)/4}
$$
are the square roots of $y$. (Why?)

If $p \equiv 1 \pmod{4}$, square roots can be found, but it is more difficult.

## How many points on an elliptic curve? {.build}

Related question: How many elements of $U(p)$ have square roots? 

Very roughly, $N \approx p$, where $N$ is the number of points on a mod $p$ elliptic curve.

More precisely:

**Hasse's Theorem.** The number of points $N$ on a mod $p$ elliptic curve satisfies the following inequality.
$$
\lvert N-p-1 \rvert < 2\sqrt{p}
$$

## Encoding messages in elliptic curves {.build}

*Problem:* To use elliptic curve groups for cryptography, we have to be able to represent messages as points on the curves.

**One Idea:** Make the message the $x$-component of a point on an elliptic curve. 

*Problem with this idea:* It will work about half the time.

**Better Idea:** Add some digits to the end of your message. Tweak them so you can find a point on the curve.

## Example

Alice wants to send the message $m = 9230923203240394234$ using a cryptosystem based on the elliptic curve $y^2 = x^3 + 7x + 9$ modulo $p = 34588345934850984359911$. 

1. Show that there is no point of the form $(m, y)$ on this elliptic curve.
2. Encode $m$ as a point on this curve by adding a digit. That is, find a point of the form $(10m+k, y)$ on this curve, for some value of $k$ between 0 and 9.
3. What is the approximate probability that this method (adding a single digit) will fail to produce a point on the curve?

## Example, part 1

Alice wants to send the message $m = 9230923203240394234$ using a cryptosystem based on the elliptic curve $y^2 = x^3 + 7x + 9$ modulo $p = 34588345934850984359911$. 

```{r}
m <- as.bigz("9230923203240394234")
p <- as.bigz("34588345934850984359911")
powm(m^3 + 7*m+ 9, (p-1)/2, p)
```

So $m^3+7m+9$ is not a perfect square. So there can't be a point $(m,y)$ on this curve.

## Example, part 2

Encode $m$ as a point on this curve by adding a digit. That is, find a point of the form $(10m+k, y)$ on this curve, for some value of $k$ between 0 and 9.

```{r}
k <- 0:9
powm((10*m+k)^3+7*(10*m+k)+9, (p-1)/2, p)
```

It looks like $k = 4$ will work.

## Example, part 2, continued

Instead of using $m = 9230923203240394234$, we use $10m + 4 = 92309232032403942344$. By the theorem on square roots, the corresponding $y$ is $\pm[(10m+4)^3+7(10m+4)+9]^{(p+1)/4}$.

```{r}
y <- powm((10*m+4)^3+7*(10*m+4)+9, (p+1)/4, p)
y^2 %% p
((10*m+4)^3+7*(10*m+4)+9) %% p
```

## How likely is this to work?

3. What is the approximate probability that this method (adding a single digit) will fail to produce a point on the curve?

