---
title: "Error-Control Codes"
date: "November 10, 2020"
output: 
  revealjs::revealjs_presentation:
    fig_width: 14 
    fig_height: 7
    self_contained: true
    theme: night
    highlight: zenburn
    css: slidesdjh.css
    center: false
    transition: slide
    reveal_options:
      controls: true
      progress: false
      width: 1080
      height: 540
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(comment = NA)
options(width = 100)
```


# Coding Theory 

## Codes

A **code** is a recipe for representing data as strings (usually binary).

- Compression: Can data be represented *efficiently* so memory/storage is conserved?
- Error Control: Can data be represented *redundantly* so errors in transmission or storage can be mitigated?

## Error Control Codes

Adversary is noise, not eavesdropper.

- Alice wants to send a message to Bob.
- In transit, parts of the message may change randomly (noisy channel).
- Bob receives the message. Can he tell what the original message was?

## Detection vs. Correction

- Error-detecting codes
    - Can we encode a message in a way that can *detect* if it was changed in transit?
- Error-correcting codes
    - Can we encode a message in a way that can *correct* any changes that may have occurred in transit?

## Error correction in English

> Dr. Hunter, I'm not felling well tocay and I won't beable to maje it to class. Soffy.

**Redundancy and Distance:** Typographical errors rarely ever obscure the meaning of English sentences, because:

- The surrounding context conveys the meaning (redundancy).
- There are no other meaningful sentences "nearby" (distance).

# Repetition Codes

## Repetition Code

Alice wants to send the message $b_1b_2\cdots b_n$ to Bob.

Instead, she chooses a number $k$ and sends the message
$$
\underbrace{b_1b_1\cdots b_1}_k \underbrace{b_2b_2\cdots b_2}_k \cdots \underbrace{b_nb_n\cdots b_n}_k
$$

## Repetition Code: Example

Alice wants to send the message $b_1b_2\cdots b_n$ to Bob.

Instead, she chooses a number $k$ and sends the message
$$
\underbrace{b_1b_1\cdots b_1}_k \underbrace{b_2b_2\cdots b_2}_k \cdots \underbrace{b_nb_n\cdots b_n}_k
$$

**Question.** Let $k=3$. Bob receives `000110111001`. What message did Alice (probably) send?

## Code Words, Binary Codes

- In a $k=3$ repetition code using the symbols `0` and `1`, the **code words** are `000` and `111`. 

- A code using only the symbols `0` and `1` is called a **binary code**.

## Repitition Code effectiveness?

[Jamboard](https://jamboard.google.com/d/1x2lBuW10YUKnf2GL2HZM9OzSjaOdBx_1QCcaGb8j36g/edit?usp=sharing) exercise: Consider a binary repetition code with $k=3$. Alice sends a 3-bit code word (`000` or `111`) to Bob.

1. Suppose Bob receives `001`. What could have happened?

2. Suppose Bob receives `001` and knows that only one error has occurred. What can he conclude?

3. Bob decides to decode a received word by finding the code word that matches the received word the closest. Suppose that the probability that a bit changes in transit is $p$. 
     - Assuming that bit-flips occur independently, calculuate (in terms of p) the probabilities of no bits changing, 1 bit changing, 2 changing, and 3 changing.
     - What is the probability that Bob's method will decode correctly?

## When is a repetition code effective?

```{r, echo=FALSE, fig.width=6, fig.height=5, fig.align='center'}
library(ggplot2)
dummy <- ggplot(data.frame(x=0), aes(x=x))
pCorrect <- function(p) {(1-p)^3+3*p*(1-p)^2}
dummy + stat_function(fun = pCorrect, size=2) + 
  xlim(0,0.5) + 
  ylab("Probability decoded correctly") + 
  xlab("Probability of bit error") +
  coord_fixed()
```

## Error correction and detection

Consider the $k=3$ repitition code.

- If we *assume* that $(\text{# of errors}) \leq 1$ in any received word, then we can always recover the word that was sent.
    - This code **corrects** one error.
- If we assume that $(\text{# of errors}) \leq 2$ in any received word, then we can always determine that an error has occurred.
    - This code **detects** two errors.
- If $(\text{# of errors}) = 3$, we won't be able to tell.

You can use the code as a *corrector* (e.g., replace `001` with `000` and continue) or as a *detector* (e.g., if you get `001`, stop and retransmit.) 


# Check Digits

## ISBN Codes

Prior to 2007, every published book was given a 10-digit ISBN number $a_1a_2a_3a_4a_5a_6a_7a_8a_9a_{10}$. Digits $a_1$ to $a_9$ are numeric, and digit $a_{10}$ could be numeric or `X`. The tenth digit $a_{10}$ is computed according to the formula
$$
a_{10} = a_1 +2a_2+3a_3+4a_4+5a_5+6a_6+7a_7+8a_8+9a_9 \bmod 11
$$
where `X` is used in place of 10.

- What if adjacent digits are transposed?
- What if a single digit is changed?

## Parity Check Digit

Alice wants to send the binary string $b_1b_2\cdots b_{n-1}$. Instead, she sends the string $b_1b_2\cdots b_{n-1}b_n$, where $b_1+b_2+\cdots+b_{n-1} = b_n$ in $\mathbb{Z}_2$.

- If Bob receives `01110110`, what can he conclude?
- If Bob receives `01110111`, what can he conclude?

## Two-Dimensional Parity Check

Alice wants to send the binary string $b_1b_2\cdots b_{16}$. Instead, she sends the following matrix.
$$
\begin{bmatrix}
b_1 & b_2 & b_3 & b_4 & a_1 \\
b_5 & b_6 & b_7 & b_8 & a_2 \\
b_9 &b_{10}&b_{11}& b_{12}& a_3 \\
b_{13} & b_{14} & b_{15} & b_{16} & a_4 \\
c_1 & c_2 & c_3 & c_4 & d
\end{bmatrix}
$$
where the $a_i$'s are parity check digits for the rows, and the $c_i$'s are parity check digits for the columns. The digit $d$ is the parity check digit for the string $c_1c_2c_3c_4$.

## Parity Check Matrix: [Jamboard](https://jamboard.google.com/d/1x2lBuW10YUKnf2GL2HZM9OzSjaOdBx_1QCcaGb8j36g/edit?usp=sharing) Exercise

Suppose Bob receives the following matrices. What can he conclude?
$$
\begin{bmatrix} 0&1&0&0&1 \\ 0&1&1&0&0 \\ 1&0&0&1&1 \\ 0&0&1&0&1 \\ 1&1&0&1&1 \end{bmatrix}
\quad\quad
\begin{bmatrix} 0&1&1&0&0 \\ 0&1&1&0&1 \\ 1&1&0&1&1 \\ 1&0&1&1&0 \\ 0&1&0&1& 0\end{bmatrix}
\quad\quad
\begin{bmatrix} 0&1&0&0&1 \\ 0&1&1&0&0 \\ 1&0&0&1&1 \\ 0&0&1&0&1 \\ 1&1&0&1& 0\end{bmatrix}
\quad\quad
$$

# Codes and Matrices

## Hamming $[7,4]$ Code

Alice wants to send the binary string $b_1b_2b_3b_4$. Instead, she sends
$$
\begin{bmatrix} b_1 & b_2 & b_3 & b_4 \end{bmatrix}
\begin{bmatrix}
1&0&0&0&1&1&0 \\
0&1&0&0&1&0&1 \\
0&0&1&0&0&1&1 \\
0&0&0&1&1&1&1
\end{bmatrix}
$$
where the entries of the matrices are computed in $\mathbb{Z}_2$.

[Jamboard Exercise.](https://jamboard.google.com/d/1x2lBuW10YUKnf2GL2HZM9OzSjaOdBx_1QCcaGb8j36g/edit?usp=sharing) How many different code words are there? Write them all down on the Jamboard.

## Hamming Distance

Let $u$ and $v$ be binary vectors. The **Hamming Distance** $d(u,v)$ is the number of positions in which $u$ and $v$ differ. That is, $d(u,v)$ is the number of 1's in $u \oplus v$.

**Question:** What is the smallest Hamming distance between any two code words in the Hamming $[7,4]$ code?

## Nearest Neighbor Decoding

In **nearest neighbor decoding**, Bob assumes that the intended message is the code word that is closest to the received message, in terms of Hamming distance.

**Example.** Use nearest neighbor decoding to decode the following messages, sent as Hamming $[7,4]$ code words.

1. `1110011`
2. `1010100`

## Linear Codes

A binary **linear code** is a binary code whose code words are formed by taking all the possible linear combinations of rows of some generating matrix. The rows of the generating matrix form a **basis** for the code.

- For example, the Hamming $[7,4]$ code is a linear code.
- The sum ($\oplus$) of two code words is another code word.
- Linear codes are vector spaces over $\mathbb{Z}_2$.

## Correction and Detection

- How many errors can a Hamming $[7,4]$ code correct?
- How many errors can a Hamming $[7,4]$ code detect?

## Hadamard Matrix

A **Hadamard matrix** is a square matrix consisting of only 1's and -1's, such that the dot product of any two rows is zero. (19th Century)

> **Exercise.** Make up a $4\times 4$ Hadamard matrix.

## Hadamard Code ([Mariner 6 and 7](https://mars.nasa.gov/mars-exploration/missions/mariner-6-7/), 1969)

Let $H$ be a $32 \times 32$ Hadamard matrix with rows $r_1, r_2, \ldots, r_{32}$. The code words are $\pm r_1, \pm r_2, \ldots, \pm r_{32}$. 

When a message $m$ is received, represent it as a column vector and calculate $Hm$.

1. What happens if $m$ is a code word?
2. What happens if $m$ has one error? Two errors?
3. How many errors can be corrected?

## Code rate

The code rate is the number of bits of information communicated divided by the number of bits in a code word.

If a binary code has $M$ code words, and the code words have length $n$, then the code rate is $\displaystyle{\frac{\log_2M}{n}}$.

> **Exercise.** Compute the code rates for 
>
> 1. The repetition code with $k=3$.
> 2. The Hamming $[7,4]$ code.
> 3. The Hadamard code.

# Matrices in R

## Defining matrices

```{r}
matrix(c(1,1,0,2,3,4, 0,1,2,1,1,4, 3,0,0,1,3,1), nrow=3, byrow=TRUE)
matrix(c(1,1,0,2,3,4, 0,1,2,1,1,4, 3,0,0,1,3,1), nrow=3, byrow=FALSE)
```

## Multiplying a row vector by a matrix

```{r}
rowVec <- c(1,2,3)
M <- matrix(c(1,1,0,2,3,4, 0,1,2,1,1,4, 3,0,0,1,3,1), nrow=3, byrow=TRUE)
M
rowVec %*% M
as.vector(rowVec %*% M)
```

## Multiplying a matrix times a column vector

```{r}
colVec <- c(1,2,3,2,1,3)
M
M %*% colVec
```

>- **Confusing:** Are vectors in R row vectors or column vectors?

## Ask the help menu

```{r, include=FALSE}
library(printr)
```

```{r}
?"%*%"
```

```{r, include=FALSE}
detach('package:printr', unload = TRUE)
```

## Conformable?

```{r, eval=FALSE}
> M
     [,1] [,2] [,3] [,4] [,5] [,6]
[1,]    1    1    0    2    3    4
[2,]    0    1    2    1    1    4
[3,]    3    0    0    1    3    1
> M %*% M
Error in M %*% M : non-conformable arguments
> 
```

## Transpose

```{r}
M
t(M)
```

## $MM^T$ and $M^TM$

```{r}
M %*% t(M)
t(M) %*% M
```


