---
title: "Codes: Definitions and Theorems"
date: "November 17, 2022"
output: 
  revealjs::revealjs_presentation:
    fig_width: 14 
    fig_height: 7
    self_contained: true
    theme: night
    highlight: zenburn
    css: slidesdjh.css
    center: false
    transition: slide
    reveal_options:
      controls: true
      progress: false
      width: 1080
      height: 540
---

# Distance in codes

## Definitions from last time

- A **binary code** $C$ is a nonempty subset of $\mathbb{Z}_2^n$. The elements of $C$ are the **code words**, and $n$ is the **length** of the code (i.e., the length of the code words).

- Let $u,v \in C$. The **Hamming distance** $d(u,v)$ is the number of positions in which $u$ and $v$ differ.

- A **nearest neighbor** to $v \in C$ is a code word $u \in C$ such that minimizes $d(u,v)$.

- The **code rate** of a binary code is $\frac{\log_2 M}{n}$, where $M$ is the number of code words and $n$ is the length of the code.

## {data-background="https://ds055uzetaobb.cloudfront.net/brioche/uploads/cxSgLAgJWR-group-1.png" data-background-size="contain"}

## The triangle inequality

**Theorem. (Triangle inequality)** Let $u,v,w$ be code words in a binary code $C$. Then
$$ d(u,v) \leq d(u,w) + d(w,v) $$

*Proof.* (Induction on the length of $C$.) 

>- Suppose the length of $C$ is 1. If $u = v$, then $d(u,v)$ = 0 and the inequality holds. If $u \neq v$, then one of $d(u,w)$ or $d(w,v)$ equals 1, and the inequality holds.

>- Suppose as inductive hypothesis that the inequality holds for codes of length $k-1$, for some $k > 1$. Let $u, v, w$ be code words of length $k$. By inductive hypothesis, the inequality holds on the first $k-1$ bits of $u,v,w$. By the above argument, it holds on the last bit. 


## Minimum distance theorem

Let $C$ be a binary code. The **minimum distance** $d(C)$ of $C$ is the minimum of $d(u,v)$ over all pairs $u,v\in C$, $u\neq v$.

> **Theorem. (Minimum distance)** Let $C$ be a binary code.
>
> 1. If $d(C) \geq s+1$, then $C$ can detect $s$ errors.
>
> 2. If $d(C) \geq 2t+1$, then $C$ can correct $t$ errors.

Proof. 1. $\checkmark$. 2. Triangle inequality.

Application: Hamming $[7,4]$ code.

## $(n,M,d)$ codes

A binary code with length $n$, $M$ code words, and minimum distance $d$ is called an $(n,M,d)$ code.

**Exercise.** Complete the table.

Code | $n$ | $M$ | $d$ | code rate | errors detected | errors corrected 
---- | --- | --- | --- | --- | --- | ---
$3$-fold repetition | $3$ |  |  | | | | 
$8$th-bit parity check | $8$ |  |  | | | |
Hamming $[7,4]$  |  |  | | | |  |
$32\times 32$ Hadamard |  |  | | | | | 

# Theoretical limits of codes

## Tradeoffs

$M =$ number of code words.

$n =$ length of the code words.

$d =$ distance of the code (min distance between words).

Tradeoffs:

- Large $M$ is good, but large $n$ is bad.
- Large $d$ is good, but generally requires larger $n$ and smaller $M$.


## Bounds on binary $(n,M,d)$ codes

>1. $M \leq 2^n$, where $M$ is the number of code words and $n$ is their length.
>2. $M \leq 2^{n-d+1}$, where $d$ is the minimum distance between code words.
>    - *Singleton bound* 
>    - Proof: Delete the first $d-1$ digits from every code word; the remaining digits still have to be different in at least one place.
>    - If $M = 2^{n-d+1}$, the code is called **maximum distance separable (MDS)**. (i.e., "$M$ acheives the Singleton bound".)

## {data-background="https://i.stack.imgur.com/HfayXm.png" data-background-size="contain"}

## Hamming Spheres

**Definition.** An $n$-dimensional **Hamming sphere** of radius $r$ with center $c$ is the set of all vectors $v$ in $\mathbb{Z}_2^n$ such that $d(v,c) \leq r$.

Exercise:

1. Write down all the vectors in a two-dimensional Hamming sphere of radius 1 centered at `00`.
2. Write down all the vectors in a three-dimensional Hamming sphere of radius 2 centered at `000`.
3. How many vectors are there in a four-dimensional Hamming sphere of radius 3 centered at `0000`?

## Number of elements in a Hamming sphere

**Lemma.** A Hamming sphere of radius $r$ in dimension $n$ has 
$$ \binom{n}{0} + \binom{n}{1} + \binom{n}{2} + \cdots + \binom{n}{r} $$
elements.

## {data-background="https://www.researchgate.net/profile/Beniamin_Bogosel/publication/317673771/figure/fig7/AS:667063202885633@1536051669021/Sphere-packing-examples-in-3D-for-problem-5.ppm" data-background-size="contain"}

## Sphere packing

- How many nonintersecting spheres can fit in a volume of size $V$?
- (# of spheres) $\times$ (volume of sphere) $\leq$ $V$

## Hamming sphere packing

>- In the space $\mathbb{Z}_2^n$ of all possible code words, how many nonintersecting Hamming spheres can fit?
>- (# of Hamming spheres) $\times$ (# of words in sphere) $\leq$ (total # of words)
>- $H \times \left[\binom{n}{0} + \binom{n}{1} + \binom{n}{2} + \cdots + \binom{n}{r} \right] \leq 2^n$, where $r$ is the radius of the sphere.

## Spheres and code distance

- Suppose a code has distance $d$. 
- Enclose each code word in a Hamming sphere of radius $t$.
- Make $t$ as large as possible.
- How are $t$ and $d$ related?

## Hamming sphere packing bound

**Theorem.** A binary $(n,M,d)$ code with $d \geq 2t+1$ satisfies
$$ M \leq \frac{2^n}{\sum_{j=0}^t \binom{n}{j}}. $$

**Definition.** A binary code with $d = 2t+1$ that achieves the above bound is called **perfect**.

That is, a perfect code is an $(n,M,d)$ code that corrects $t$ errors such that $$M = \frac{2^n}{\sum_{j=0}^t \binom{n}{j}}.$$ 

Is the Hamming $(7,4)$ code perfect?

# Linear Codes

## {data-background="https://upload.wikimedia.org/wikipedia/commons/7/72/Vector_addition_ans_scaling.png" data-background-size="contain" data-background-color="white"}

## Just-in-time linear algebra

- Vector Space
- Basis
- Generating set/matrix of a code

## Linear Codes

A binary **linear code** is a binary code whose code words are formed by taking all the possible linear combinations of rows of some generating matrix.

- Linear codes are vector spaces over $\mathbb{Z}_2$.
- The rows of the generating matrix form a **basis** for the code.
    - For example, the Hamming $[7,4]$ code is a linear code.
- The sum ($\oplus$) of two code words is another code word.

# R tips and tricks

## Defining matrices

```{r}
matrix(c(1,1,0,2,3,4, 0,1,2,1,1,4, 3,0,0,1,3,1), nrow=3, byrow=TRUE)
matrix(c(1,1,0,2,3,4, 0,1,2,1,1,4, 3,0,0,1,3,1), nrow=3, byrow=FALSE)
```

## Multiplying a row vector by a matrix

```{r}
rowVec <- c(1,2,3)
M <- matrix(c(1,1,0,2,3,4, 0,1,2,1,1,4, 3,0,0,1,3,1), nrow=3, byrow=TRUE)
M
rowVec %*% M
as.vector(rowVec %*% M)
```

## Multiplying a matrix times a column vector

```{r}
colVec <- c(1,2,3,2,1,3)
M
M %*% colVec
```

>- **Confusing:** Are vectors in R row vectors or column vectors?

## Ask the help menu

```{r, include=FALSE}
library(printr)
```

```{r}
?"%*%"
```

```{r, include=FALSE}
detach('package:printr', unload = TRUE)
```

## Conformable?

```{r, eval=FALSE}
> M
     [,1] [,2] [,3] [,4] [,5] [,6]
[1,]    1    1    0    2    3    4
[2,]    0    1    2    1    1    4
[3,]    3    0    0    1    3    1
> M %*% M
Error in M %*% M : non-conformable arguments
> 
```

## Transpose

```{r}
M
t(M)
```

## $MM^T$ and $M^TM$

```{r}
M %*% t(M)
t(M) %*% M
```

## Converting numbers to binary vectors 

For the assignment, you might want to be able to convert an integer `n` to a binary vector `v` (e.g., if you want to represent all possible binary vectors of a certain length.)

Notice that `intToBits` gives you 32 bits, no matter what. 

```{r}
intToBits(13)
```

Make those numeric instead of binary, so matrix multiplication will work:

```{r}
as.numeric(intToBits(13))
```

## Subscript the digits you want

You probably don't want all those 0's, so you can just subscript the ones you want.

```{r}
as.numeric(intToBits(13))[1:4]
```

The result is the binary vector corresponding to the binary representation of 13, with the digits reversed. If you want them in the correct order, you can reverse the subscripts:

```{r}
as.numeric(intToBits(13))[4:1]
```



