---
title: "Cyclic Codes"
date: "November 19, 2020"
output: 
  revealjs::revealjs_presentation:
    fig_width: 14 
    fig_height: 7
    self_contained: true
    theme: night
    highlight: zenburn
    css: slidesdjh.css
    center: false
    transition: slide
    reveal_options:
      controls: true
      progress: false
      width: 1080
      height: 540
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(comment = NA)
options(width = 100)
```

# Golay Codes

## {data-background="https://upload.wikimedia.org/wikipedia/commons/thumb/6/60/Voyager_spacecraft_model.png/1280px-Voyager_spacecraft_model.png" data-background-size="contain"}

## Golay $[24,12]$-code (Voyager I, II, 1979)

The Golay code $G_{24}$ has the following generator matrix.

```{r, results='hide'}
golay24 <- matrix(c(1,0,0,0,0,0,0,0,0,0,0,0,1,1,1,0,1,1,1,0,0,0,1,0,
                    0,1,0,0,0,0,0,0,0,0,0,0,1,0,1,1,0,1,1,1,0,0,0,1,
                    0,0,1,0,0,0,0,0,0,0,0,0,1,1,0,1,1,0,1,1,1,0,0,0,
                    0,0,0,1,0,0,0,0,0,0,0,0,1,0,1,0,1,1,0,1,1,1,0,0,
                    0,0,0,0,1,0,0,0,0,0,0,0,1,0,0,1,0,1,1,0,1,1,1,0,
                    0,0,0,0,0,1,0,0,0,0,0,0,1,0,0,0,1,0,1,1,0,1,1,1,
                    0,0,0,0,0,0,1,0,0,0,0,0,1,1,0,0,0,1,0,1,1,0,1,1,
                    0,0,0,0,0,0,0,1,0,0,0,0,1,1,1,0,0,0,1,0,1,1,0,1,
                    0,0,0,0,0,0,0,0,1,0,0,0,1,1,1,1,0,0,0,1,0,1,1,0,
                    0,0,0,0,0,0,0,0,0,1,0,0,1,0,1,1,1,0,0,0,1,0,1,1,
                    0,0,0,0,0,0,0,0,0,0,1,0,1,1,0,1,1,1,0,0,0,1,0,1,
                    0,0,0,0,0,0,0,0,0,0,0,1,0,1,1,1,1,1,1,1,1,1,1,1),
                  byrow=TRUE, nrow=12)
```

## Decoding options

In order of speed (slow to fast):

- Search for nearest neighbor.
- Compute coset of received word $r$, then use coset leader.
- Compute syndrome of received word $r$, then use table.

## Coset Leader Syndrome lookup table 

```{r, eval=FALSE}
cosetLeaderSyndromeTable <- function(G) {
  n <- ncol(G)
  k <- nrow(G)
  C <- generateCode(G) # TODO (or done)
  H <- parityCheckMatrix(G) # TODO
  Z2n <- t(sapply(0:(2^n-1), function(x){as.numeric(intToBits(x)[1:n])}))
  allsynd <- sapply(1:2^n, function(x) {syndrome(Z2n[x,],H)}) # TODO (syndrome)
  uniqueReps <- which(!duplicated(allsynd))
  syndromes <- allsynd[uniqueReps]
  cosetLeaders <- lapply(uniqueReps, function(x) {cosetLeader(Z2n[x,],C)} ) # TODO (cosetLeader)
  names(cosetLeaders) <- syndromes
  return(cosetLeaders)
}
```

## Golay lookup table

```{r}
load("golay24table.RData") # made using cosetLeaderSyndromeTable
gT$`010100001111`
gT[["010100001111"]]
head(gT,10) # first 10 entries
```


## Golay $[24,12]$ decoding shortcut

The Golay code is *self-dual*, meaning its generator matrix can function as a parity check matrix.

Bob receives $r = c+e$. He computes the syndrome $s = rG^T$. (Note: this is a different way of computing the syndrome.)

>1. If $\mbox{wt}(s) \leq 3$, then $e = [s \,\,\, \mathbf{0}]$.
>2. Let $B$ be such that $G = [I \, B]$. If $\mbox{wt}(sB)\leq 3$, then $e = [\mathbf{0}\,\,\, sB]$
>3. Let $c_{13}, c_{14}, \ldots, c_{24}$ be the columns of $B$. If $\mbox{wt}(s+c_j^T) \leq 2$ for some $j$, then $e = [s+c_j^T \,\,\, \mathbf{0}] +u_j$, where $u_j$ has a 1 in the $j$th position and 0's elsewhere.
>4. Let $b_1, b_2, \ldots, b_{12}$ be the columns of $B^T$. If $\mbox{wt}(sB+b_j^T) \leq 2$ for some $j$, then $e = [\mathbf{0} \,\,\, sB+b_j^T] + u_j$. 

## Golay $[24,12]$ decoding shortcut example 

Bob receives $m = c+e$. He computes the "syndrome" $s = rG^T$. 

```{r}
mess <- c(1,1,1,0,0,0,1,0,1,0,0,0,0,1,0,1,0,0,0,0,1,0,1,1)
s <- (mess %*% t(golay24) ) %% 2
s
```

*1.* If $\mbox{wt}(s) \leq 3$, then $e = [s \,\,\, \mathbf{0}]$.

## Golay $[24,12]$ decoding shortcut example 

*2.* Let $B$ be such that $G = [I \, B]$. If $\mbox{wt}(sB)\leq 3$, then $e = [\mathbf{0}\,\,\, sB]$

```{r}
B <- golay24[,13:24]
sB <- (s %*% B) %% 2
sB
```

## Golay $[24,12]$ decoding shortcut example {.smaller}

*3.* Let $c_{13}, c_{14}, \ldots, c_{24}$ be the columns of $B$. If $\mbox{wt}(s+c_j^T) \leq 2$, then $e = [s+c_j^T \,\,\, \mathbf{0}] +u_j$, where $u_j$ has a 1 in the $j$th position and 0's elsewhere.

```{r}
sapply(1:12, function(j) {(s + B[,j]) %% 2})
```

So $j = 18$, and $e = \scriptsize{\begin{bmatrix} 0 &0 &0 &0 &0 &0 &0 &1 &0 &0 &0 &0 &0 &0 & 0 &0 &0 &1 &0 &0 &0 \end{bmatrix}}$.

# Cyclic Codes

## More structure $\Rightarrow$ Faster decoding

- For a general code, finding the nearest neighbor of a length $n$ code word is *NP-complete.*
- Linear codes allow for syndrome decoding, which is faster than searching for the nearest neighbor.
- The Golay codes have additional constraints allowing for faster decoding.
- Other codes with more algebraic structure also have faster decoding algorithms.

## Cyclic codes

**Definition.** A code is **cyclic** if it is closed under cyclic shift:
$$
(c_1, c_2, \ldots, c_{n-1}, c_n) \in C \Rightarrow (c_n, c_1, c_2, \ldots, c_{n-1}) \in C
$$

Examples: 8th bit parity check, $k$-fold repetition code

[Jamboard](https://jamboard.google.com/d/1hQg7Df49pdmeGBT1V2TC3cl7j7VkgQ8_HshFVoQkG8U/edit?usp=sharing) Exercise: For each given code word, find the smallest cyclic linear code containing the code word.

1. $\begin{bmatrix} 1 & 0 &  0 & 1 \end{bmatrix}$

2. $\begin{bmatrix} 1 & 1 & 0 & 0 & 1 & 0 & 1 \end{bmatrix}$

## Code words and polynomials

Let's start indexing at zero.

- A bit string $(c_0, c_1, \ldots, c_{n-1})$ corresponds to a polynomial $c_0 +c_1x + \cdots + c_{n-1}x^{n-1}$.
- Adding polynomials corresponds to adding bit strings in $\mathbb{Z}_2^n$.
- How about shifting?

## Quotient rings

>- Recall that the ring $\mathbb{Z}_2[x]/(x^n+1)$ consists of all polynomials in $x$ under the "mod" relation that $x^n + 1 = 0$, i.e., $x^n = 1$.
>- The elements of $\mathbb{Z}_2[x]/(x^n+1)$ are the polynomials $c_0 +c_1x + \cdots + c_{n-1}x^{n-1}$, which correspond to the $2^n$ bit strings of length $n$.
- Notice: 
$$ \scriptsize
\begin{align}
x(c_0 +c_1x + \cdots + c_{n-1}x^{n-1}) &= c_0x +c_1x^2 + \cdots + c_{n-2}x^{n-1} + c_{n-1}x^{n} \\ &= c_0x +c_1x^2 + \cdots + c_{n-2}x^{n-1} + c_{n-1} \\ &= c_{n-1} + c_0x +c_1x^2 + \cdots + c_{n-2}x^{n-1} 
\end{align}
$$
- So *multiplication by $x$ is the same as shifting.*

## Cyclic Codes are Ideals

**Definition.** A cyclic linear code (i.e., a **cyclic code**) of length $n$ is a subset of $\mathbb{Z}_2[x]/(x^n+1)$ that is closed under addition and closed under multiplication by $x$.

- (By the distributive property, closure under multiplication by $x$ is the same as closure under multiplication by any polynomial.)
- In algebra, subsets of rings with these closure properties are called **ideals**.

## Fact from Algebra

**Fact:** Every ideal $C$ of $\mathbb{Z}_2[x]/(x^n+1)$ has the form
$$
C = \{p(x)g(x) \mid p(x) \in \mathbb{Z}_2[x]/(x^n+1) \}
$$
for some *generating polynomial* $g(x)$ (which is the polynomial of least degree in $C$).

- In this case, we write $C = \langle g(x) \rangle$, the ideal generated by $g(x)$.
- [Jamboard](https://jamboard.google.com/d/1hQg7Df49pdmeGBT1V2TC3cl7j7VkgQ8_HshFVoQkG8U/edit?usp=sharing) Exercise, continued: Find the generating polynomials for these two cyclic codes.

## More Algebra Facts

- If $\gcd(g(x), x^n + 1) = 1$, then $\langle g(x) \rangle$ is the whole ring (all the bit strings).
- The factors of $x^n + 1$, and their products, generate all the codes.
- Example: Investigate factors of $x^7 + 1$.

## Tool: Macaulay2

```
i1 : R = ZZ/2[x]
o1 = R
o1 : PolynomialRing

i2 : factor(x^7+1)
              3           3    2
o2 = (x + 1)(x  + x + 1)(x  + x  + 1)
o2 : Expression of class Product
```

Online version: http://web.macaulay2.com/

## Characterization of Cyclic Codes

>- **Theorem.** There is a one-to-one correspondence between the divisors of $x^n + 1$ and the cyclic codes of length $n$.
>- **Theorem.** If $g(x)$ is a divisor of $x^n + 1$ of degree $r$, then the code words of $\langle g(x) \rangle$ are uniquely determined by products of the form $g(x)p(x)$, where $p(x)$ is a polynomial of degree less than $n-r$.
>- **Corollary.** If $g(x)$ is a divisor of $x^n + 1$ of degree $r$, then $\langle g(x) \rangle$ is an $[n, n-r]$ cyclic code.
>    - The smaller the polynomial, the bigger the code.

## Parity check polynomials 

Let $C = \langle g(x) \rangle$ be a cyclic code. If $m(x)$ is a polynomial satisfying $m(x)g(x) = x^n+1$, then $m(x)$ is called a **parity check polynomial.** 

>- **Theorem.** If $m(x)$ is a parity check polynomial and $p(x)$ is a message, then
$m(x)p(x) = 0$ in $\mathbb{Z}_2[x]/(x^n+1)$
if and only if $p(x)$ is a codeword.

## Cyclic Code Examples

Over $\mathbb{Z}_2$, $x^7+1 = (x+1)(x^3+x+1)(x^3+x^2+1)$.

| $g(x)$ | $[n,k]$ | Code |
| ------ | ------- | ---- |
| $x+1$  | $[7,6]$  | parity check |
| $x^3 + x + 1$ or $x^3 + x^2 + 1$ | $[7,4]$ | Hamming |
| $(x+1)(x^3 + x + 1)$ or $(x+1)(x^3 + x^2 + 1)$ | $[7,3]$ | Hamming dual |
| $(x^3 + x + 1)(x^3 + x^2 + 1) = x^6+x^5+x^4+x^3+x^2+x+1$ | $[7,1]$ | repetition |

