---
title: "Cyclic Codes"
date: "November 29, 2022"
output: 
  revealjs::revealjs_presentation:
    fig_width: 14 
    fig_height: 7
    self_contained: true
    theme: night
    highlight: zenburn
    css: slidesdjh.css
    center: false
    transition: slide
    reveal_options:
      controls: true
      progress: false
      width: 1080
      height: 540
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(comment = NA)
options(width = 100)
```

# Last Time: Linear Codes

## Cosets and Syndromes

Suppose $C$ is a linear code, and a code word $w$ is sent, and $r$ is received.

- If you can find the *coset* $r + C$, then its coset leader $e$ gives you the nearest neighbor $r+e$.
- You can compute the *syndrome* $rH^T$, and this can help you find the coset leader (given a table).

## Minimum distance of Hamming codes

**Lemma.** Let $C$ be a systematic linear binary code, and let $H$ be its parity check matrix.

1. If $C$ contains a codeword of weight 1, then $H$ contains a zero column.
2. If $C$ contains a codeword of weight 2, then $H$ has two identical columns.

. . .

**Corollary:** The minimum distance of a Hamming code is 3, and they're perfect.

## Hamming codes are perfect

**Corollary:** The minimum distance of a Hamming code is 3, and they're perfect.

Proof: Hamming codes are constructed by forming $H$ first, where $H$ has $2^m-1$ different nonzero columns, so the minimum distance is greater than 2, by the Lemma. The code has length $n = 2^m-1$, and $M = 2^k$ code words, where $k = 2^m-1-m$. Observe:

$$
\frac{2^n}{\binom{n}{0} + \binom{n}{1}} = \frac{2^n}{1+n} = \frac{2^{2^m-1}}{2^m} = 2^k = M
$$

Therefore $d > 3$ is impossible, because $M$ would then exceed the Hamming bound. So $d = 3$, and the code is perfect. 

## Hamming Syndrome Decoding

- The nonzero syndromes are the columns of the parity check matrix $H$. Proof:
    - Hamming codes can correct 1 error.
        - If $r$ has one error, then $r = w+\delta_i$, where $w \in C$.
        - $\delta_i$ is the vector with a 1 in the $i$th position, 0s elsewhere.
    - Syndrome of $r$ is $rH^T = wH^T + \delta_iH^T = \delta_iH^T$.
        - This accounts for $n+1 = 2^m$ syndromes (including $\mathbf{0}$).
        - The cosets have size $M = 2^k$. So this accounts for $2^m\cdot 2^k = 2^n$ possible bit strings (i.e. all of them). So these are *all* the syndromes.
- For nearest neighbor decoding, the column of the syndrome is the location of the error. Proof:
    - For any coset, the coset leader must be $\delta_i$.

## Example: The Hamming (7,4) code


```{r}
m <- 3
powersOf2 <- 2^(0:(m-1))
hamH <- sapply(c(setdiff(1:(2^m-1), powersOf2), powersOf2), 
       function(x){as.numeric(intToBits(x)[1:m])})
hamH
```

Correct the error in $(1,0,1,0,1,1,1)$

```{r}
(c(1,0,1,0,1,1,1) %*% t(hamH)) %% 2
```

The error is in position 6: $w = (1,0,1,0,1,0,1)$


# Cyclic Codes

## More structure $\Rightarrow$ Faster decoding

- For a general code, finding the nearest neighbor of a length $n$ code word is *NP-complete.*
- Linear codes allow for syndrome decoding, which is faster than searching for the nearest neighbor.
- The Golay codes have additional constraints allowing for faster decoding.
- Other codes with more algebraic structure also have faster decoding algorithms.

## Cyclic codes

**Definition.** A code is **cyclic** if it is closed under cyclic shift:
$$
(c_1, c_2, \ldots, c_{n-1}, c_n) \in C \Rightarrow (c_n, c_1, c_2, \ldots, c_{n-1}) \in C
$$

Examples: 8th bit parity check, $k$-fold repetition code

Exercise: For each given code word, find the smallest cyclic linear code containing the code word.

1. $\begin{bmatrix} 1 & 0 &  0 & 1 \end{bmatrix}$

2. $\begin{bmatrix} 1 & 1 & 0 & 0 & 1 & 0 & 1 \end{bmatrix}$

## Code words and polynomials

Let's start indexing at zero.

- A bit string $(c_0, c_1, \ldots, c_{n-1})$ corresponds to a polynomial $c_0 +c_1x + \cdots + c_{n-1}x^{n-1}$.
- Adding polynomials corresponds to adding bit strings in $\mathbb{Z}_2^n$.
- How about shifting?

## Quotient rings

>- Recall that the ring $\mathbb{Z}_2[x]/(x^n+1)$ consists of all polynomials in $x$ under the "mod" relation that $x^n + 1 = 0$, i.e., $x^n = 1$.
>- The elements of $\mathbb{Z}_2[x]/(x^n+1)$ are the polynomials $c_0 +c_1x + \cdots + c_{n-1}x^{n-1}$, which correspond to the $2^n$ bit strings of length $n$.
- Notice: 
$$ \scriptsize
\begin{align}
x(c_0 +c_1x + \cdots + c_{n-1}x^{n-1}) &= c_0x +c_1x^2 + \cdots + c_{n-2}x^{n-1} + c_{n-1}x^{n} \\ &= c_0x +c_1x^2 + \cdots + c_{n-2}x^{n-1} + c_{n-1} \\ &= c_{n-1} + c_0x +c_1x^2 + \cdots + c_{n-2}x^{n-1} 
\end{align}
$$
- So *multiplication by $x$ is the same as shifting.*

## Cyclic Codes are Ideals

**Definition.** A cyclic linear code (i.e., a **cyclic code**) of length $n$ is a subset of $\mathbb{Z}_2[x]/(x^n+1)$ that is closed under addition and closed under multiplication by $x$.

- (By the distributive property, closure under multiplication by $x$ is the same as closure under multiplication by any polynomial.)
- In algebra, subsets of rings with these closure properties are called **ideals**.

## Fact from Algebra

**Fact:** Every ideal $C$ of $\mathbb{Z}_2[x]/(x^n+1)$ has the form
$$
C = \{p(x)g(x) \mid p(x) \in \mathbb{Z}_2[x]/(x^n+1) \}
$$
for some *generating polynomial* $g(x)$ (which is the polynomial of least degree in $C$).

- In this case, we write $C = \langle g(x) \rangle$, the ideal generated by $g(x)$.
- Exercise, continued: Find the generating polynomials for these two cyclic codes.

## More Algebra Facts

- If $\gcd(g(x), x^n + 1) = 1$, then $\langle g(x) \rangle$ is the whole ring (all the bit strings).
- The factors of $x^n + 1$, and their products, generate all the codes.
- Example: Investigate factors of $x^7 + 1$.

## Tool: Macaulay2

```
i1 : R = ZZ/2[x]
o1 = R
o1 : PolynomialRing

i2 : factor(x^7+1)
              3           3    2
o2 = (x + 1)(x  + x + 1)(x  + x  + 1)
o2 : Expression of class Product
```

Online version: http://web.macaulay2.com/

## Characterization of Cyclic Codes

>- **Theorem.** There is a one-to-one correspondence between the divisors of $x^n + 1$ and the cyclic codes of length $n$.
>- **Theorem.** If $g(x)$ is a divisor of $x^n + 1$ of degree $r$, then the code words of $\langle g(x) \rangle$ are uniquely determined by products of the form $g(x)p(x)$, where $p(x)$ is a polynomial of degree less than $n-r$.
>- **Corollary.** If $g(x)$ is a divisor of $x^n + 1$ of degree $r$, then $\langle g(x) \rangle$ is an $[n, n-r]$ cyclic code.
>    - The smaller the polynomial, the bigger the code.

## Parity check polynomials 

Let $C = \langle g(x) \rangle$ be a cyclic code. If $m(x)$ is a polynomial satisfying $m(x)g(x) = x^n+1$, then $m(x)$ is called a **parity check polynomial.** 

>- **Theorem.** If $m(x)$ is a parity check polynomial and $p(x)$ is a message, then
$m(x)p(x) = 0$ in $\mathbb{Z}_2[x]/(x^n+1)$
if and only if $p(x)$ is a codeword.

## Cyclic Code Examples

Over $\mathbb{Z}_2$, $x^7+1 = (x+1)(x^3+x+1)(x^3+x^2+1)$.

| $g(x)$ | $[n,k]$ | Code |
| ------ | ------- | ---- |
| $x+1$  | $[7,6]$  | parity check |
| $x^3 + x + 1$ or $x^3 + x^2 + 1$ | $[7,4]$ | Hamming |
| $(x+1)(x^3 + x + 1)$ or $(x+1)(x^3 + x^2 + 1)$ | $[7,3]$ | Hamming dual |
| $(x^3 + x + 1)(x^3 + x^2 + 1) = x^6+x^5+x^4+x^3+x^2+x+1$ | $[7,1]$ | repetition |

