---
title: "BCH Codes"
date: "December 1, 2022"
output: 
  revealjs::revealjs_presentation:
    fig_width: 14 
    fig_height: 7
    self_contained: true
    theme: night
    highlight: zenburn
    css: slidesdjh.css
    center: false
    transition: slide
    reveal_options:
      controls: true
      progress: false
      width: 1080
      height: 540
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(comment = NA)
options(width = 100)
```

# Recall: Cyclic codes

## Cyclic Code Definitions

 **Definition.** A code is **cyclic** if it is closed under cyclic shift:
$$
(c_1, c_2, \ldots, c_{n-1}, c_n) \in C \Rightarrow (c_n, c_1, c_2, \ldots, c_{n-1}) \in C
$$

**Definition.** A cyclic linear code (i.e., a **cyclic code**) of length $n$ is a subset of $\mathbb{Z}_2[x]/(x^n+1)$ that is closed under addition and closed under multiplication by $x$. (A cyclic code is an *ideal* of $\mathbb{Z}_2[x]/(x^n+1)$.)

## Cyclic Code Facts

>- There is a one-to-one correspondence between the divisors of $x^n + 1$ and the cyclic codes of length $n$.
>- If $p(x)$ is a divisor of $x^n + 1$ of degree $r$, then the code words of $\langle p(x) \rangle$ are uniquely determined by products of the form $p(x)q(x)$, where $q(x)$ is a polynomial of degree less than $n-r$.
>- If $p(x)$ is a divisor of $x^n + 1$ of degree $r$, then $\langle p(x) \rangle$ is a $[n, n-r]$ cyclic code.
>- Let $\langle g(x) \rangle$ be the principal ideal corresponding to a cyclic code $C$. If $m(x)$ is a polynomial satisfying $m(x)g(x) = x^n+1$, then $m(x)$ is called a **parity check polynomial.** If $m(x)$ is a parity check polynomial and $p(x)$ is a message, then $m(x)p(x) = 0$ in $\mathbb{Z}_2[x]/(x^n+1)$ if and only if $p(x)$ is a codeword.

## Cyclic Code Examples

Over $\mathbb{Z}_2$, $x^7+1 = (x+1)(x^3+x+1)(x^3+x^2+1)$.

| $p(x)$ | $[n,k]$ | Code |
| ------ | ------- | ---- |
| $x+1$  | $[7,6]$  | parity check |
| $x^3 + x + 1$ or $x^3 + x^2 + 1$ | $[7,4]$ | Hamming |
| $(x+1)(x^3 + x + 1)$ or $(x+1)(x^3 + x^2 + 1)$ | $[7,3]$ | Hamming dual |
| $(x^3 + x + 1)(x^3 + x^2 + 1) = x^6+x^5+x^4+x^3+x^2+x+1$ | $[7,1]$ | repetition |

# Fancier Factoring

## Motivating questions

- In high school algebra, can you factor $x^2 + 4x + 3$?
- In high school algebra, can you factor $x^2 + 4x + 5$?

## Review: Finite Fields

>- If $p(x)$ is an irreducible polynomial of degree $n$ in $\mathbb{Z}_2[x]$, then $\mathbb{Z}_2[x]/(p(x))$ is a field with $2^n$ elements, called $\mbox{GF}(2^n)$.
>- We think of the elements of $\mbox{GF}(2^n)$ as polynomials, under the modulus relation $p(x) = 0$.
>- Equivalently, we can think of $\mbox{GF}(2^n)$ as a field containing $\mathbb{Z}_2$ and an element $\alpha$, such that $p(\alpha) = 0$. We say that $\mbox{GF}(2^n)$ is an *extension field* of $\mathbb{Z}_2$. **Notation:** $\mbox{GF}(2^n) = \mathbb{Z}_2(\alpha)$
>    - Analogy: $\mathbb{C}$ is an extension field of $\mathbb{R}$, containing $i$, a root of $x^2 + 1$.
>    - $\mathbb{C} = \mathbb{R}(i)$

## Roots of polynomials

$\mbox{GF}(2^n)$ is a field containing $\mathbb{Z}_2$ and an element $\alpha$, such that $p(\alpha) = 0$. We say that $\mbox{GF}(2^n)$ is an *extension field* of $\mathbb{Z}_2$.

Example: $\mbox{GF(4)} = \mathbb{Z}_2[x]/(x^2+x+1)$

- Suppose $\alpha$ is a root of $x^2+x+1$.
- Then $\alpha^2 + \alpha + 1 = 0$.
- The powers of alpha are
    - $\alpha^1 = \alpha$
    - $\alpha^2 = \alpha + 1$
    - $\alpha^3 = \alpha^2 + \alpha = 1$
    
## Exploration: Powers of a root

1. Over $\mathbb{Z}_2$, let $\alpha$ be a root of $x^3 + x + 1$. Compute all the powers of $\alpha$, simplifying when possible. How many are there?

2. Over $\mathbb{Z}_2$, let $\alpha$ be a root of $x^3 + x^2 + x + 1$. Compute all the powers of $\alpha$, simplifying when possible. How many are there?

## Primitive roots of unity

**Facts:** $\mbox{GF}(2^n)$ contains an element $\alpha$ such that:

- $\alpha^{2^n-1} = 1$. ($\alpha$ is a $(2^n-1)$th *root of unity*)
- Every nonzero element of $\mbox{GF}(2^n)$ is a power of $\alpha$. ($\alpha$ is *primitive*)
- $\alpha$ is a root of a polynomial of degree $n$. (a *primitive polynomial*)

Examples: $\mbox{GF(4)}$, $\mbox{GF(8)}$

[List of primitive polynomials](https://www.jjj.de/mathdata/all-primpoly.txt)

## Factoring $x^7 + 1$ over $\mathbb{Z}_2$

```
i1 : R = ZZ/2[x]

o1 = R

o1 : PolynomialRing

i2 : factor(x^7+1)

              3           3    2
o2 = (x + 1)(x  + x + 1)(x  + x  + 1)

o2 : Expression of class Product

```

## Factoring $x^7 + 1$ over $\mbox{GF}(8)$

```
i1 : R = GF(8)[x]

o1 = R

o1 : PolynomialRing

i2 : factor(x^7+1)

                         2                  2           2               2
o2 = (x + 1)(x + a)(x + a )(x + a + 1)(x + a  + a)(x + a  + a + 1)(x + a  + 1)

o2 : Expression of class Product
```

## Which primitive polynomial is [Macaulay2](https://www.unimelb-macaulay2.cloud.edu.au/) using?

```
i1 : F = GF(8)
o1 = F
o1 : GaloisField

i2 : ambient F
        ZZ
        --[a]
         2
o2 = ----------
      3
     a  + a + 1
o2 : QuotientRing

i3 : apply(1..7, i -> a^i)
          2          2       2           2
o3 = (a, a , a + 1, a  + a, a  + a + 1, a  + 1, 1)
o3 : Sequence
```

## Splitting $x^n + 1$ 

**Fact:** Let $\alpha$ be a primitive $n$th root of unity in some extension field of $\mathbb{Z}_2$. Then 

$$x^n + 1 = (x+1)(x+\alpha)(x+\alpha^2)\cdots(x+\alpha^{n-1})$$

So a divisor $g(x)$ of $x^n+1$ (and therefore a cyclic code $\langle g(x) \rangle$) determines a set of powers of $\alpha$ that are zeroes of $g(x)$.

- **Jargon:** $x^n+1$ *splits* over $\mathbb{Z}_2(\alpha)$.
- Analogy: All real polynomials split over $\mathbb{C} = \mathbb{R}(i)$.

## Group Exercise

Let $C=\langle g(x)\rangle$ be a cyclic code in $\mathbb{Z}_2[x]/(x^{13}+1)$, and let $\alpha$ be a primitive $13$th root of unity. Suppose that $g(\alpha^4) = g(\alpha^5) = 0$.

1. Let $p(x)$ be the parity check polynomial such that $g(x)p(x) = x^{13} + 1$. Use the fact that $x^{13}+1$ splits over $\mathbb{Z}_2(\alpha)$ to explain why $p(\alpha^4) \neq 0 \neq p(\alpha^5)$.

2. Suppose that, for some $k$, the polynomial $m(x) = x^k$ is a code word of weight 1. Show that this leads to a contradiction.

3. Suppose that, for some $j,k$, the polynomial $m(x) = x^j+ x^k$ is a code word of weight 2. Show that this leads to a contradiction.

4. Explain why the minimum distance of $C$ is at least 3.

# BCH Codes

## BCH Bound Theorem over $\mathbb{Z}_2$

**Theorem.** Let $C = \langle g(x) \rangle$ be a cyclic code in $\mathbb{Z}_2[x]/(x^n+1)$, for $n$ odd. Let $\alpha$ be a primitive $n$th root of unity, and suppose that for some integers $l$ and $\delta$, $$g(\alpha^l) = g(\alpha^{l+1}) = \cdots = g(\alpha^{l+\delta})=0.$$  Then the minimum distance of $C$ is at least $\delta + 2$.

## BCH Code of designed distance $d$

Let $\alpha$ be a primitive $n$th root of unity.

- Factor $x^n+1$ over $\mathbb{Z}_2$: $x^n+1 = f_1(x)f_2(x)\cdots f_r(x)$.
- Write each factor $f_i(x)$ as a product of $(x + \alpha^j)$'s.
- Choose $g(x)$ to be a product of factors $f_i(x)$ containing $\delta+1$ consecutive $(x+\alpha^j)$'s.
- By the theorem, $\langle g(x) \rangle$ will have distance at least $\delta + 2$.

## Examples

Macaulay2 tells us that over $\mathbb{Z}_2$,

$$x^7+1 = (x+1)(x^3+x^2+1)(x^3+x+1)$$

Furthermore, over $\text{GF}(8)$, 

$$
\begin{align}
x^3+x^2+1 &= (x + \alpha^3)(x+\alpha^5)(x+\alpha^6) \\
x^3+x+1 &= (x+\alpha)(x+\alpha^2)(x+\alpha^4)
\end{align}
$$

Exercise. In $\mathbb{Z}_2[x]/(x^7+1)$,

1. Find a generator for a code of designed distance 3.
2. Find a generator for a code of designed distance 4.
3. Find a generator for a code of designed distance 7.

## BCH Decoding

- Error-detection of BCH codes is easy, because they are cyclic codes.
- BCH codes have [good error-correcting decoding algorithms](https://infoscience.epfl.ch/record/33884/files/HongV95.pdf).
- To illustrate, we'll see how to correct a single error.

## Correcting 1 error of a BCH code

Let $\langle g(x) \rangle$ be a BCH code in $\mathbb{Z}_2[x]/(x^n+1)$ of designed distance $d \geq 3$, so $g(\alpha^l) = g(\alpha^{l+1})=0$ for some $l$ and for a primitive $n$th root of unity $\alpha$.  

Let $r(x) = m(x) + e(x)$, where $m(x)$ is a code word and $e(x)$ is an error vector.

1. Compute $s = r(\alpha^l)$. If $s = 0$, there is no error, or there is more than one error. (why?) 
2. If $s\neq 0$, compute $s^{-1} r(\alpha^{l+1})$. This will be a power of $\alpha$ (why?), and the power (plus 1) gives the location of the error.

## Why it works

Let $\langle g(x) \rangle$ be a BCH code in $\mathbb{Z}_2[x]/(x^n+1)$ of designed distance $d \geq 3$, so $g(\alpha^l) = g(\alpha^{l+1})=0$ for some $l$ and for a primitive $n$th root of unity $\alpha$.  Let $r(x) = m(x) + e(x)$, where $m(x)$ is a code word and $e(x)$ is an error vector.

1. Compute $s = r(\alpha^l)$. If $s = 0$, there is no error, or there is more than one error. 

\begin{align}
s &= r(\alpha^l) \\
  &= m(\alpha^l) + e(\alpha^l) \\
  &= 0 + e(\alpha^l) && \mbox{ since } m(x) \in \langle g(x) \rangle
\end{align}

## Why it works

2. If $s\neq 0$, compute $s^{-1} r(\alpha^{l+1})$. This will be a power of $\alpha$, and the power (plus 1) gives the location of the error. 

Suppose there is one error, so $e(x) = x^k$ for some $k$.

\begin{align}
s^{-1}r(\alpha^{l+1}) &= (e(\alpha^l))^{-1}e(\alpha^{l+1}) && \mbox{ by part 1} \\
&= ((\alpha^l)^k)^{-1}(\alpha^{l+1})^k \\
&= \alpha^k && \mbox{ properties of exponents}
\end{align}


# Reed-Solomon Codes

## {data-background="https://solarsystem.nasa.gov/system/resources/detail_files/2304_mgs_1280.jpg" data-background-size="contain"}

## Reed-Solomon Codes and Mars

The Mars Global Surveyor (MGS) was launched in 1996 and retired in 2007. According to [this journal article](https://onlinelibrary.wiley.com/doi/abs/10.1002/sat.871) by a JPL scientist,

> The MGS downlink to Earth incorporates a 25 W travelling wave tube amplifier and a gimballed, 1:5 m high-gain antenna with a transmit gain of 39:1 dBic. Forward error correction coding options include a $(255, 223)$ Reed-Solomon code concatenated with either a $(7, \frac{1}{2})$ or $(15, \frac{1}{6})$ convolutional code.

## Codes over other fields

- So far, we have been working with codes over $\mathbb{Z}_2$.
- We can also consider codes over other finite fields.
- *Hamming distance* between two code words is still the *number of positions where the words differ*.
- For linear codes, the weight (= number of nonzero entries) of the lightest code word is still the minimum distance of the code.

## BCH Bound Theorem over $\mathbb{Z}_2$

Recall,

**Theorem.** Let $C = \langle g(x) \rangle$ be a cyclic code in $\mathbb{Z}_2[x]/(x^n+1)$, for $n$ odd. Let $\alpha$ be a primitive $n$th root of unity, and suppose that for some integers $l$ and $\delta$, $$g(\alpha^l) = g(\alpha^{l+1}) = \cdots = g(\alpha^{l+\delta})=0.$$  Then the minimum distance of $C$ is at least $\delta + 2$.

## BCH Bound Theorem over any field $F$

The BCH Bound Theorem generalizes to other finite fields.

**Theorem.** Let $C = \langle g(x) \rangle$ be a cyclic code in $F[x]/(x^n-1)$, where $F$ has $q=p^m$ elements. Assume $p \nmid n$. Let $\alpha$ be a primitive $n$th root of unity, and suppose that for some integers $l$ and $\delta$, $$g(\alpha^l) = g(\alpha^{l+1}) = \cdots = g(\alpha^{l+\delta})=0.$$  Then the minimum distance of $C$ is at least $\delta + 2$.

## Reed-Solomon Codes

**Fact:** Suppose $F$ is finite field with $q$ elements. If $n = q-1$, then $F$ contains a primitive $n$th root of unity $\alpha$.

So we can construct a code in $F[x]/(x^n-1)$ using a generating polynomial of the following form.

$$
g(x) = (x-\alpha)(x-\alpha^2)\cdots(x-\alpha^{d-1})
$$

and then the theorem tells us the minimum distance of $\langle g(x) \rangle$ is at least $d$.

Such $\langle g(x) \rangle$ is called a *Reed-Solomon code*.

## Reed-Solomon optimality

- A code in $F[x]/(x^n-1)$ with generating polynomial
$$
g(x) = (x-\alpha)(x-\alpha^2)\cdots(x-\alpha^{d-1})
$$
has minimum distance at least $d$. But $g(x)$ has only $d$ nonzero terms. So the minimum distance *equals* $d$.
- If $|F|=q$, then there are $q^{n-(d-1)}$ code words, since $g(x)$ has degree $d-1$.
- But $q^{n-d+1}$ is the Singleton bound. So these codes are *Maximum Distance Separable* (MDS).

## Reed-Solomon Example: $F = \mathbb{Z}_7$

Consider $F = \mathbb{Z}_7$. A Reed-Solomon code is an ideal of $\mathbb{Z}_7[x]/(x^6-1)$.

We need a primitive 6th root of unity in $\mathbb{Z}_7$.

```{r}
outer(1:6, 1:6, function(x,y) {(x^y) %% 7})
```

## Reed-Solomon Example: $F = \mathbb{Z}_7$ {.build}

Consider $F = \mathbb{Z}_7$. A Reed-Solomon code is an ideal of $\mathbb{Z}_7[x]/(x^6-1)$.

Let $\alpha = 3$ be our primitive 6th root of unity in $\mathbb{Z}_7$.

A distance 4 code is generated by $g(x) = (x-3)(x-3^2)(x-3^3) = x^3+3x^2+x+6$.

The dimension of this code is $k = n - \deg(g(x)) = 6-3 = 3$. Therefore a generating matrix is:

$$
G = \begin{bmatrix} 6 & 1 & 3 & 1 & 0 & 0 \\
0 & 6 & 1 & 3 & 1 & 0  \\
0 & 0 & 6 & 1 & 3 & 1  
\end{bmatrix}
$$

The number of code words is $7^3 = 343$.

## Reed-Solomon Example: $F = \text{GF}(8)$

- Recall that $\alpha$ is a primitive 7th root of unity satisfying $\alpha^3+\alpha+1 = 0$
- $\text{GF}(8) = \{0, 1, \alpha, \alpha^2, \ldots, \alpha^6\}$ 
- A Reed-Solomon code is an ideal of $\text{GF}(8)[x]/(x^7+1)$. 
- A Reed-Solomon code can be generated by $g(x) = (x+\alpha)(x+\alpha^2) \stackrel{?}{=} x^2 + \alpha^4x + \alpha^3$.

## Reed-Solomon Example: $F = \text{GF}(8)$

Exercise. Let $g(x) = x^2 + \alpha^4x + \alpha^3$ be the generator for a Reed-Solomon code in $\text{GF}(8)[x]/(x^7+1)$. 

1. What is the minimum distance of $\langle g(x) \rangle$?
2. What is the length of this code?
3. What is the dimension of this code?
4. Write down a generating matrix for this code.
5. Find a code word that's not in your generating matrix.
6. How many code words are there in $\langle g(x) \rangle$?

## Correcting 1 error of a Reed-Solomon code

Let $\langle g(x) \rangle$ be a Reed-Solomon code in $F[x]/(x^n-1)$ of distance $d \geq 3$, so $g(\alpha) = g(\alpha^{2})=0$ for a primitive $n$th root of unity $\alpha$, where $n+1$ is the size of $F$.  

Let $r(x) = m(x) + e(x)$, where $m(x)$ is a code word and $e(x)$ is an error vector.

1. Compute $s = r(\alpha)$. If $s = 0$, there is no error, or there is more than one error.
2. If $s \neq 0$, compute $s^{-1} r(\alpha^{2})$. This will have the form $\alpha^k$, and error will be in the $x^k$ term.
3. The error vector is $e(x) = s\alpha^{-k}x^k$ (why?), so the decoded word is $r(x) - e(x)$.

## Example {.smaller}

For the code $\langle x^2 + \alpha^4x + \alpha^3 \rangle$ in $\text{GF}(8)[x]/(x^7+1)$, decode the received word
$$
r(x) = x^6 + \alpha^3x^4 + \alpha^2x^3 + \alpha^4x^2 + \alpha^6 x + \alpha^6
$$

```
i1 : R = GF(8)[x]
o1 = R
o1 : PolynomialRing

i2 : r = x^6 + a^3*x^4 + a^2*x^3 + a^4*x^2 + a^6*x + a^6
      6           4    2 3     2      2     2          2
o2 = x  + (a + 1)x  + a x  + (a  + a)x  + (a  + 1)x + a  + 1
o2 : R

i3 : s = sub(r, x=>a)
o3 = a
o3 : GF 8
```

So $s = \alpha$. Continuing, we get  $s^{-1} r(\alpha^{2}) = \alpha^3$, so the error is in the $x^3$ term. The error vector is $e(x) = s\alpha^{-3}x^3 = \alpha^{-2}x^3 = \alpha^5x^3$. The decoded word is $r(x)+e(x) = x^6 + \alpha^3x^4 + \alpha^3x^3 + \alpha^4x^2 + \alpha^6 x + \alpha^6$.

## $\text{GF}(2^m)$ and bit strings

Recall that the elements of $\text{GF}(2^m)$ can be regarded as bit strings. For example,

```
i1 : F = GF(8)
o1 = F
o1 : GaloisField


i3 : apply(1..7, i -> a^i)
          2          2       2           2
o3 = (a, a , a + 1, a  + a, a  + a + 1, a  + 1, 1)
o3 : Sequence
```

We can identify these with 010, 001, 110, 011, 111, 101, 100, and along with 000 we get all the 3-digit bit strings.


## Mars Global Surveyor Code

The MGS used a $(255, 233)$ Reed-Solomon code.

- What is $F$?
- What is the generator?
- What is the minimum distance? How many errors can be corrected?

## Good for finding corrupted *Bytes*

- An element of $\text{GF}(2^8)$ is an 8-bit string, i.e., a byte.
- A Reed-Solomon code over $\text{GF}(256)$ can correct corrupted *bytes*.
- If errors occur in consecutive bits (e. g.,  a scratch on a CD or a solar flare), this is a good thing.
- Not so good if errors are randomly distributed.

