---
title: "Reed-Solomon Codes"
date: "December 3, 2020"
output: 
  revealjs::revealjs_presentation:
    fig_width: 14 
    fig_height: 7
    self_contained: true
    theme: night
    highlight: zenburn
    css: slidesdjh.css
    center: false
    transition: slide
    reveal_options:
      controls: true
      progress: false
      width: 1080
      height: 540
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(comment = NA)
options(width = 100)
```

# Equivalence of Codes

## Equality

- Two codes are *equal* if they contain the same code words.
- Example: `G` of Assignment #19 and $\langle x^{10}+x^8+x^5+x^4+x^2+x+1 \rangle$

## Permutation Equivalence

- Two codes are *permutation equivalent* if they are equal after permuting the coordinates of one of them.
- In other words, the code words (as matrices) have the same columns, except ordered differently.
- Permutation equivalent codes have the same distance properties, but possibly not the same algebraic properties.

## Example: Hamming $[7,3]$

Systematic generating matrix:

```
gnrtCd> generateCode(matrix(c(1,0,0,0,1,1,0,
gnrtCd+                       0,1,0,0,1,0,1,
gnrtCd+                       0,0,1,0,0,1,1,
gnrtCd+                       0,0,0,1,1,1,1), nrow=4, byrow=TRUE))
      [,1] [,2] [,3] [,4] [,5] [,6] [,7]
 [1,]    0    0    0    0    0    0    0
 [2,]    1    0    0    0    1    1    0
 [3,]    0    1    0    0    1    0    1
 [4,]    1    1    0    0    0    1    1
 [5,]    0    0    1    0    0    1    1
 [6,]    1    0    1    0    1    0    1
 [7,]    0    1    1    0    1    1    0
 [8,]    1    1    1    0    0    0    0
 [9,]    0    0    0    1    1    1    1
[10,]    1    0    0    1    0    0    1
[11,]    0    1    0    1    0    1    0
[12,]    1    1    0    1    1    0    0
[13,]    0    0    1    1    1    0    0
[14,]    1    0    1    1    0    1    0
[15,]    0    1    1    1    0    0    1
[16,]    1    1    1    1    1    1    1
```

## Hamming $[7,3]$ as cyclic code

Not quite the same code:

```
> generateCode(cyclicCodeGenMat(c(3,1,0), 7))
      [,1] [,2] [,3] [,4] [,5] [,6] [,7]
 [1,]    0    0    0    0    0    0    0
 [2,]    1    1    0    1    0    0    0
 [3,]    0    1    1    0    1    0    0
 [4,]    1    0    1    1    1    0    0
 [5,]    0    0    1    1    0    1    0
 [6,]    1    1    1    0    0    1    0
 [7,]    0    1    0    1    1    1    0
 [8,]    1    0    0    0    1    1    0
 [9,]    0    0    0    1    1    0    1
[10,]    1    1    0    0    1    0    1
[11,]    0    1    1    1    0    0    1
[12,]    1    0    1    0    0    0    1
[13,]    0    0    1    0    1    1    1
[14,]    1    1    1    1    1    1    1
[15,]    0    1    0    0    0    1    1
[16,]    1    0    0    1    0    1    1
```

## Permutation matrix

We want to start with the identity matrix and swap rows 3 and 4 and also rows 5 and 6:

```{r}
M <- diag(7) # identity
M[, c(3,4)] <- M[, c(4,3)] # swap columns 3 and 4
M[, c(5,6)] <- M[, c(6,5)] # swap columns 5 and 6
M
```

## Permutation equivalent code

Permute the code (or the generating matrix), and the codes match.

```
generateCode(cyclicCodeGenMat(c(3,1,0), 7) %*% M)
      [,1] [,2] [,3] [,4] [,5] [,6] [,7]
 [1,]    0    0    0    0    0    0    0
 [2,]    1    1    1    0    0    0    0
 [3,]    0    1    0    1    0    1    0
 [4,]    1    0    1    1    0    1    0
 [5,]    0    0    1    1    1    0    0
 [6,]    1    1    0    1    1    0    0
 [7,]    0    1    1    0    1    1    0
 [8,]    1    0    0    0    1    1    0
 [9,]    0    0    1    0    0    1    1
[10,]    1    1    0    0    0    1    1
[11,]    0    1    1    1    0    0    1
[12,]    1    0    0    1    0    0    1
[13,]    0    0    0    1    1    1    1
[14,]    1    1    1    1    1    1    1
[15,]    0    1    0    0    1    0    1
[16,]    1    0    1    0    1    0    1
```

## Permute columns by multiplying by $M$ on right

```
> M
     [,1] [,2] [,3] [,4] [,5] [,6] [,7]
[1,]    1    0    0    0    0    0    0
[2,]    0    1    0    0    0    0    0
[3,]    0    0    0    1    0    0    0
[4,]    0    0    1    0    0    0    0
[5,]    0    0    0    0    0    1    0
[6,]    0    0    0    0    1    0    0
[7,]    0    0    0    0    0    0    1
> cyclicCodeGenMat(c(3,1,0), 7)
     [,1] [,2] [,3] [,4] [,5] [,6] [,7]
[1,]    1    1    0    1    0    0    0
[2,]    0    1    1    0    1    0    0
[3,]    0    0    1    1    0    1    0
[4,]    0    0    0    1    1    0    1
> cyclicCodeGenMat(c(3,1,0), 7) %*% M
     [,1] [,2] [,3] [,4] [,5] [,6] [,7]
[1,]    1    1    1    0    0    0    0
[2,]    0    1    0    1    0    1    0
[3,]    0    0    1    1    1    0    0
[4,]    0    0    1    0    0    1    1
```

# Cyclic Codes so far

## Recall Cyclic codes, BCH codes

- The code words of a cyclic code are all multiples of a generating polynomial, under the modulus $x^n+1 = 0$.
- The generating polynomial of a BCH code contains factors (in a splitting field) of the form $(x-\alpha^i), (x-\alpha^{i+1}), \ldots (x-\alpha^{i+\delta})$, where $\alpha$ is a primitive $n$th root of unity.
- Such a code is guaranteed to have minimum distance at least $\delta + 2$ (i.e., "designed distance").

## BCH Decoding

- Error-detection of BCH codes is easy, because they are cyclic codes.
- BCH codes have [good error-correcting decoding algorithms](https://infoscience.epfl.ch/record/33884/files/HongV95.pdf).
- To illustrate, we'll see how to correct a single error.

## Correcting 1 error of a BCH code

Let $\langle g(x) \rangle$ be a BCH code in $\mathbb{Z}_2[x]/(x^n+1)$ of designed distance $d \geq 3$, so $g(\alpha^l) = g(\alpha^{l+1})=0$ for some $l$ and for a primitive $n$th root of unity $\alpha$.  

Let $r(x) = m(x) + e(x)$, where $m(x)$ is a code word and $e(x)$ is an error vector.

1. Compute $s = r(\alpha^l)$. If $s = 0$, there is no error, or there is more than one error. (why?) 
2. If $s\neq 0$, compute $s^{-1} r(\alpha^{l+1})$. This will be a power of $\alpha$ (why?), and the power (plus 1) gives the location of the error.

## Why it works

Let $\langle g(x) \rangle$ be a BCH code in $\mathbb{Z}_2[x]/(x^n+1)$ of designed distance $d \geq 3$, so $g(\alpha^l) = g(\alpha^{l+1})=0$ for some $l$ and for a primitive $n$th root of unity $\alpha$.  Let $r(x) = m(x) + e(x)$, where $m(x)$ is a code word and $e(x)$ is an error vector.

1. Compute $s = r(\alpha^l)$. If $s = 0$, there is no error, or there is more than one error. 

\begin{align}
s &= r(\alpha^l) \\
  &= m(\alpha^l) + e(\alpha^l) \\
  &= 0 + e(\alpha^l) && \mbox{ since } m(x) \in \langle g(x) \rangle
\end{align}

## Why it works

2. If $s\neq 0$, compute $s^{-1} r(\alpha^{l+1})$. This will be a power of $\alpha$, and the power (plus 1) gives the location of the error. 

Suppose there is one error, so $e(x) = x^k$ for some $k$.

\begin{align}
s^{-1}r(\alpha^{l+1}) &= (e(\alpha^l))^{-1}e(\alpha^{l+1}) && \mbox{ by part 1} \\
&= ((\alpha^l)^k)^{-1}(\alpha^{l+1})^k \\
&= \alpha^k && \mbox{ properties of exponents}
\end{align}

(See also pp. 438-439.)

## BCH Error Correction [Exercise](https://jamboard.google.com/d/1g_67Zxk30YU1d9LZqxlG6ur9YYG8ky4rl9N7vEK6h0A/edit?usp=sharing)

Recall that $x^3+x^2+1 = (x + \alpha^3)(x+\alpha^5)(x+\alpha^6)$ generates a $[7, 4, 3]$ code in $\mathbb{Z}_2[x]/(x^7+1)$. Here $\text{GF}(8) = \mathbb{Z}_2[x]/(x^3+x+1) = \mathbb{Z}_2(\alpha)$.  Decode $x^6 + x^4 +x$, if possible. Use [Macaulay2](http://web.macaulay2.com) appropriately. 

```
i1 : F = GF(8)
o1 = F
o1 : GaloisField

i2 : ambient F
        ZZ
        --[a]
         2
o2 = ----------
      3
     a  + a + 1
o2 : QuotientRing

i3 : apply(1..7, i -> a^i)
          2          2       2           2
o3 = (a, a , a + 1, a  + a, a  + a + 1, a  + 1, 1)
o3 : Sequence
```

# Reed-Solomon Codes

## {data-background="https://solarsystem.nasa.gov/system/resources/detail_files/2304_mgs_1280.jpg" data-background-size="contain"}

## Reed-Solomon Codes and Mars

The Mars Global Surveyor (MGS) was launched in 1996 and retired in 2007. According to [this journal article](https://onlinelibrary.wiley.com/doi/abs/10.1002/sat.871) by a JPL scientist,

> The MGS downlink to Earth incorporates a 25 W travelling wave tube amplifier and a gimballed, 1:5 m high-gain antenna with a transmit gain of 39:1 dBic. Forward error correction coding options include a $(255, 223)$ Reed-Solomon code concatenated with either a $(7, \frac{1}{2})$ or $(15, \frac{1}{6})$ convolutional code.

## Codes over other fields

- So far, we have been working with codes over $\mathbb{Z}_2$.
- We can also consider codes over other finite fields.
- *Hamming distance* between two code words is still the *number of positions where the words differ*.
- For linear codes, the weight (= number of nonzero entries) of the lightest code word is still the minimum distance of the code.

## BCH Bound Theorem over $\mathbb{Z}_2$

Recall,

**Theorem.** Let $C = \langle g(x) \rangle$ be a cyclic code in $\mathbb{Z}_2[x]/(x^n+1)$, for $n$ odd. Let $\alpha$ be a primitive $n$th root of unity, and suppose that for some integers $l$ and $\delta$, $$g(\alpha^l) = g(\alpha^{l+1}) = \cdots = g(\alpha^{l+\delta})=0.$$  Then the minimum distance of $C$ is at least $\delta + 2$.

## BCH Bound Theorem over any field $F$

The BCH Bound Theorem generalizes to other finite fields.

**Theorem.** Let $C = \langle g(x) \rangle$ be a cyclic code in $F[x]/(x^n-1)$, where $F$ has $q=p^m$ elements. Assume $p \nmid n$. Let $\alpha$ be a primitive $n$th root of unity, and suppose that for some integers $l$ and $\delta$, $$g(\alpha^l) = g(\alpha^{l+1}) = \cdots = g(\alpha^{l+\delta})=0.$$  Then the minimum distance of $C$ is at least $\delta + 2$.

## Reed-Solomon Codes

**Fact:** Suppose $F$ is finite field with $q$ elements. If $n = q-1$, then $F$ contains a primitive $n$th root of unity $\alpha$.

So we can construct a code in $F[x]/(x^n-1)$ using a generating polynomial of the following form.

$$
g(x) = (x-\alpha)(x-\alpha^2)\cdots(x-\alpha^{d-1})
$$

and then the theorem tells us the minimum distance of $\langle g(x) \rangle$ is at least $d$.

Such $\langle g(x) \rangle$ is called a *Reed-Solomon code*.

## Reed-Solomon optimality

- A code in $F[x]/(x^n-1)$ with generating polynomial
$$
g(x) = (x-\alpha)(x-\alpha^2)\cdots(x-\alpha^{d-1})
$$
has minimum distance at least $d$. But $g(x)$ has only $d$ nonzero terms. So the minimum distance *equals* $d$.
- If $|F|=q$, then there are $q^{n-(d-1)}$ code words, since $g(x)$ has degree $d-1$.
- But $q^{n-d+1}$ is the Singleton bound. So these codes are *Maximum Distance Separable* (MDS).

## Reed-Solomon Example: $F = \mathbb{Z}_7$

Consider $F = \mathbb{Z}_7$. A Reed-Solomon code is an ideal of $\mathbb{Z}_7[x]/(x^6-1)$.

We need a primitive 6th root of unity in $\mathbb{Z}_7$.

```{r}
outer(1:6, 1:6, function(x,y) {(x^y) %% 7})
```

## Reed-Solomon Example: $F = \mathbb{Z}_7$ {.build}

Consider $F = \mathbb{Z}_7$. A Reed-Solomon code is an ideal of $\mathbb{Z}_7[x]/(x^6-1)$.

Let $\alpha = 3$ be our primitive 6th root of unity in $\mathbb{Z}_7$.

A distance 4 code is generated by $g(x) = (x-3)(x-3^2)(x-3^3) = x^3+3x^2+x+6$.

The dimension of this code is $k = n - \deg(g(x)) = 6-3 = 3$. Therefore a generating matrix is:

$$
G = \begin{bmatrix} 6 & 1 & 3 & 1 & 0 & 0 \\
0 & 6 & 1 & 3 & 1 & 0  \\
0 & 0 & 6 & 1 & 3 & 1  
\end{bmatrix}
$$

The number of code words is $7^3 = 343$.

## Reed-Solomon Example: $F = \text{GF}(8)$

- Recall that $\alpha$ is a primitive 7th root of unity satisfying $\alpha^3+\alpha+1 = 0$
- $\text{GF}(8) = \{0, 1, \alpha, \alpha^2, \ldots, \alpha^6\}$ 
- A Reed-Solomon code is an ideal of $\text{GF}(8)[x]/(x^7+1)$. 
- A Reed-Solomon code can be generated by $g(x) = (x+\alpha)(x+\alpha^2) \stackrel{?}{=} x^2 + \alpha^4x + \alpha^3$.

## Reed-Solomon Example: $F = \text{GF}(8)$

[Exercise](https://jamboard.google.com/d/13LkM9xzzWSoTHn6aISwcov9_7_7n10isGQWDpCn5_Hc/edit?usp=sharing). Let $g(x) = x^2 + \alpha^4x + \alpha^3$ be the generator for a Reed-Solomon code in $\text{GF}(8)[x]/(x^7+1)$. 

1. What is the minimum distance of $\langle g(x) \rangle$?
2. What is the length of this code?
3. What is the dimension of this code?
4. Write down a generating matrix for this code.
5. Find a code word that's not in your generating matrix.
6. How many code words are there in $\langle g(x) \rangle$?

## Correcting 1 error of a Reed-Solomon code

Let $\langle g(x) \rangle$ be a Reed-Solomon code in $F[x]/(x^n-1)$ of distance $d \geq 3$, so $g(\alpha) = g(\alpha^{2})=0$ for a primitive $n$th root of unity $\alpha$, where $n+1$ is the size of $F$.  

Let $r(x) = m(x) + e(x)$, where $m(x)$ is a code word and $e(x)$ is an error vector.

1. Compute $s = r(\alpha)$. If $s = 0$, there is no error, or there is more than one error.
2. If $s \neq 0$, compute $s^{-1} r(\alpha^{2})$. This will have the form $\alpha^k$, and error will be in the $x^k$ term.
3. The error vector is $e(x) = s\alpha^{-k}x^k$ (why?), so the decoded word is $r(x) - e(x)$.

## Example {.smaller}

For the code $\langle x^2 + \alpha^4x + \alpha^3 \rangle$ in $\text{GF}(8)[x]/(x^7+1)$, decode the received word
$$
r(x) = x^6 + \alpha^3x^4 + \alpha^2x^3 + \alpha^4x^2 + \alpha^6 x + \alpha^6
$$

```
i1 : R = GF(8)[x]
o1 = R
o1 : PolynomialRing

i2 : r = x^6 + a^3*x^4 + a^2*x^3 + a^4*x^2 + a^6*x + a^6
      6           4    2 3     2      2     2          2
o2 = x  + (a + 1)x  + a x  + (a  + a)x  + (a  + 1)x + a  + 1
o2 : R

i3 : s = sub(r, x=>a)
o3 = a
o3 : GF 8
```

So $s = \alpha$. Continuing, we get  $s^{-1} r(\alpha^{2}) = \alpha^3$, so the error is in the $x^3$ term. The error vector is $e(x) = s\alpha^{-3}x^3 = \alpha^{-2}x^3 = \alpha^5x^3$. The decoded word is $r(x)+e(x) = x^6 + \alpha^3x^4 + \alpha^3x^3 + \alpha^4x^2 + \alpha^6 x + \alpha^6$.

## $\text{GF}(2^m)$ and bit strings

Recall that the elements of $\text{GF}(2^m)$ can be regarded as bit strings. For example,

```
i1 : F = GF(8)
o1 = F
o1 : GaloisField


i3 : apply(1..7, i -> a^i)
          2          2       2           2
o3 = (a, a , a + 1, a  + a, a  + a + 1, a  + 1, 1)
o3 : Sequence
```

We can identify these with 010, 001, 110, 011, 111, 101, 100, and along with 000 we get all the 3-digit bit strings.


## Mars Global Surveyor Code

The MGS used a $(255, 233)$ Reed-Solomon code.

- What is $F$?
- What is the generator?
- What is the minimum distance? How many errors can be corrected?

## Good for finding corrupted *Bytes*

- An element of $\text{GF}(2^8)$ is an 8-bit string, i.e., a byte.
- A Reed-Solomon code over $\text{GF}(256)$ can correct corrupted *bytes*.
- If errors occur in consecutive bits (e. g.,  a scratch on a CD or a solar flare), this is a good thing.
- Not so good if errors are randomly distributed.

