---
geometry: left=1in, right=0.5in, top=0.5in, bottom=0.5in
classoption: twoside
output: 
  pdf_document:
    includes:
      in_header: activityheader.tex
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(plot3D)
```
## Activity: Elliptic Curve Experiments I \hfill Name:\underline{\mbox{}\hspace{2.5in}} 

The graphs below show three elliptic curves $y^2 = x^3+5x+10$, $y^2=x^3+10$, and $y^2=x^3-5x$.  For each curve $E$, determine if each statement below is true or false.

1. If $(x,y) \in E$, so is $(x,-y)$.
2. If $(x,y) \in E$, so is $(-x, y)$.
3. If $P$ and $Q$ are points on $E$, then the line through $P$ and $Q$ intersects $E$ in a third point $R$. (Try some examples with a straightedge.)
4. There is a line that intersects $E$ in four points. (See if you can find one using a straightedge.)
5. The infinite component of $E$ has two inflection points.
6. There is more than one line that intersects the curve in only one point. (Can you find more than one?)


```{r, fig.width=8, fig.height=7}
par(mfrow=c(1,3))
x<-seq(-5,10,length=1000)
y<-seq(-30,30,length=1000)
z<-outer(x,y,function(x,y) -y^2 + x^3+5*x ) 
contour(x,y,z,levels=0, labels="", labcex=0.1, asp=1)

x<-seq(-5,10,length=1000)
y<-seq(-30,30,length=1000)
z<-outer(x,y,function(x,y) -y^2 + x^3+10 ) 
contour(x,y,z,levels=0, labels="", labcex=0.1, asp=1)

x<-seq(-5,10,length=1000)
y<-seq(-30,30,length=1000)
z<-outer(x,y,function(x,y) -y^2 + x^3-5*x  ) 
contour(x,y,z,levels=0, labels="", labcex=0.1, asp=1)
```

