---
title: "Tools for Breaking a Vigenère Cipher"
output:
  html_document:
    toc: false
    df_print: paged
    theme: spacelab
---

## Assignment: due Sunday, 11:59 pm

1. Add the functions `letterCounts`, `digramTable`, `vigenere`, `shiftVec`, `skipString`, and `letterFreq` to your R package, with appropriate documentation and examples as `roxygen` comments. Also, fix any issues that you may have had with Version 0.1.0. Check to make sure you aren't getting any warning messages. If you are getting a warning message about the `NAMESPACE` file, delete the `NAMESPACE` file, and then when you Install and Restart and ROxygen will build a new `NAMESPACE` file for you. Make sure that all of the documentation gets built properly.

2. Write a function `findVigKey` that consumes a ciphertext string `ciphertext` and a key length `keyLength`, and returns the key that produced the ciphertext, assuming that the ciphertext was produced using a Vigenère cipher with key length `keyLength`. Most/all of the code you need for this is in the slides, you just need to piece it together. Here's a skeleton:
```{r, eval=FALSE}
findVigKey <- function(ciphertext, keyLength) {
  vKey <- numeric(keyLength) # preallocate a vector of the desired length
  englishFreqs <- c(0.082,0.015,0.028,0.043,0.127,0.022,0.020,0.061,0.070, 0.002,0.008,0.040,0.024,
                  0.067,0.075,0.019,0.001,0.060,0.063,0.091,0.028,0.010,0.023,0.001,0.020,0.001)
  # Insert code here to find the key, and assign the pth element of the key (as an element
  # of Z_26) the variable vKey[p], for values of p from 1 to keyLength
  return(vKey)
}
```

Test your `findVigKey` on the following example. Make it one of the examples in your documentation.

```{r, eval=FALSE}
ptext <- "howeverinthisextremedistresshewasnotdestituteofhisusualsagacitybuttrustinghimselftotheprovidenceofgodheputhislifeintohazardinthemannerfollowingandnowsaidhesinceitisresolvedamongyouthatyouwilldie"
keyAsVector <- stringToMod26("skyler")
ctext <- vigenere(ptext, keyAsVector)
findVigKey(ctext, 6)  # These...
keyAsVector           #    should match.
```
Add this function to your package, with the appropriate documentation and examples.

 3. Edit the `DESCRIPTION` file and update the version number to `0.1.1`. After you have built and reloaded your package, tested the examples, and checked the help files, click on `More` in the build menu, and build a source package. A new file called `cryptoXYZ_0.1.1.tar.gz` should appear in the parent directory. Upload this file to Canvas.
 
