---
title: "Discrete Logarithms and Hash Functions"
output:
  html_document:
    toc: false
    df_print: paged
    theme: spacelab
---

## Assignment: due Sunday, 11:59 pm

Add the file [`dlhash.R`](https://djhunter.github.io/cryptography/R/dlhash.R) to the `R` subdirectory of your project. 

0. This file includes a function `miniSHA` that imports the `openssl` package (which you need to install if you haven't already.) Unfortunately, there is a naming collision with the `digest` package that produces annoying warning messages. To make these messages go away, we need to fix `aestools.R` so that it only imports the function `AES` and not any other functions. To do this, change the line in `aestools.R` from
```{r, eval=FALSE}
#' @import digest
```
to
```{r, eval=FALSE}
#' @importFrom digest AES
```

1. Implement the function `nextPrimRoot` as described in the documentation. This will automate the process that you worked through in the last written assignment.

2. Implement `discreteLogBrute` as specified. Your function should just try all the possible powers of `alpha` until it finds one that equals `beta`.

3. Implement `discreteLogBSGS` as specified. There is some advice in the slides about using the `match` function.

4. Implement `birthdayAttack` as specified. There is some advice in the slides about using the `duplicated` function.
