---
title: "Hash Functions and DSA"
output:
  html_document:
    toc: false
    df_print: paged
    theme: spacelab
---

## Assignment: due Sunday, 11:59 pm

Add the file [`hashdsa.R`](https://djhunter.github.io/cryptography/R/hashdsa.R) to the `R` subdirectory of your project. 

0. This file includes a function `miniSHA` that imports the `openssl` package (which you need to install if you haven't already.) Unfortunately, there is a naming collision with the `digest` package that produces annoying warning messages. To make these messages go away, we need to fix `aestools.R` so that it only imports the function `AES` and not any other functions. To do this, change the line in `aestools.R` from
```{r, eval=FALSE}
#' @import digest
```
to
```{r, eval=FALSE}
#' @importFrom digest AES
```

1. Implement `birthdayAttack` as specified. There is some advice in the slides about using the `duplicated` function.

2. Implement the function `setUpDSA` as described in the documentation. This function should generate the DSA parameters $q$ and $p$ such that $q \mid (p-1)$, $q$ is an $N$-bit prime, and $p$ is an $L$-bit prime. (Hint: find $q$ first, then search for a suitable $p$. If you use `urand.bigz`, make sure it is giving you the right size number.)