---
title: "Mod p Elliptic Curves"
output:
  html_document:
    toc: false
    df_print: paged
    theme: spacelab
---

## Assignment: due Sunday, 11:59 pm

Add the file [`modp_elliptic.R`](https://djhunter.github.io/cryptography/R/modp_elliptic.R) to the `R` subdirectory of your project. 

1. Implement the function `ecAddModp` as described in the documentation. Since this function is intended to be used for Lenstra's algorithm, we need it to print a warning if either of the inverse calculation fails (which should sometimes happen if the modulus is not prime). See `?warning` for details on how to print a warning.

2. Implement `ecNegModp` as specified. 

3. Implement `ecPowModp` as specified. This needs to be an elliptic curve version of the `qPower` function [from the notes](../slides/15elliptic_curves2.html#/efficient-exponentiation), using repeated "squaring" (doubling). Otherwise it will be too slow. It only needs to work for nonnegative exponents. The given examples should run quickly:
```{r, eval=FALSE}
> system.time(ecPowModp(3, 45, 8831, c(4,11), as.bigz("2349089023472938409283490823")))
   user  system elapsed 
  0.033   0.000   0.033
```

4. Implement `ecDiscreteLog` as specified, using the [baby-step, giant-step algorithm](../slides/16elliptic_curves3.html#/attacking-ec-discrete-logarithms-bsgs). You may find that `list`s are faster than matrices for storing elliptic curve points, i.e., pairs of `bigz`'s. It should be able to do the given examples in a couple seconds. The "challenge" example might take a couple minutes. See if you can do better than this:
```{r, eval=FALSE}
> system.time(ecDiscreteLog(4, -12063, 34543427, c(23,14), c(10735908, 411234)))
   user  system elapsed 
  1.486   0.000   1.489 
> system.time(ecDiscreteLog(4, 1, as.bigz("426904703359"), c(4,9),
                            as.bigz(c("26917083261", "24329765219")),
                            N=500000))
   user  system elapsed 
160.866   0.216 161.243 
```