---
title: "Prime numbers and RSA"
output:
  html_document:
    toc: false
    df_print: paged
    theme: spacelab
---

Upload to Canvas a PDF of your work for the following problems. 

## Assignment: due Monday, 11:59 pm

1. Use your Fermat primality test `FermatTest` to find all of the weak pseudoprimes between 1000 and 10000 for the base 7. Use the `isprime` function to determine that your pseudoprimes are composite.

2. Consider the following (composite) number.

```{r, eval=FALSE}
bigComp <- as.bigz("41495155688809929585124078636911611510124462322424
                   368999956573296906528114129081463997081530206029675
                   796566441088639070653584409712811422453347196222375
                   13761725653531896260925123613")
```

Check that this number is composite by using your Fermat test. Notice that the square of the following number is close to `bigComp`.  

```{r, eval=FALSE}
sqRoot <- as.bigz("203703597633448608626844568840937816105146839366593
                  6250636140449354381299763336706183397647")
```

Use this information, and the fact that the factors of `bigComp` are approximately the same size, to factor `bigComp` using the Fermat factorization method ([T] p. 182). 
   
3. Suppose Buford implements RSA with the following choices.

```{r, eval=FALSE}
p <- as.bigz(39241)
q <- as.bigz(95347)
e <- as.bigz(9007)
```

Compute his decryption exponent $d$, and try encrypting the plaintext 1000, and then decrypting the result. What happens? Which part of the [derivation](https://djhunter.github.io/cryptography/slides/09rsa1.html#/proof-that-rsa-works-how-to-recover-m) fails? Why does it fail?

