---
title: "Systematic Codes"
output:
  html_document:
    toc: false
    df_print: paged
    theme: spacelab
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(comment = NA)
```

Upload to Canvas a PDF of your work for the following problems. 

## Assignment: due Wednesday, 11:59 pm

1. Use the fact that $vH^T = \mathbf{0}$ for any codeword $v$ to prove both parts of the following lemma.

> **Lemma.** Let $C$ be a sytematic linear binary code, and let $H$ be its parity check matrix.
>
> 1. If $C$ contains a codeword of weight 1, then $H$ contains a zero column.
> 2. If $C$ contains a codeword of weight 2, then $H$ has two identical columns.

2. The following questions refer to the generating matrix `G` and message `mess` given below.
    a. The matrix `G` is a generating matrix for a systematic linear binary code. How many codewords are there in this code?
    b. How many errors can this code correct?
    c. How many different syndromes are there?
    d. Compute the parity check matrix $H$.
    e. Bob receives the message `mess`. Find its syndrome.
    f. Determine the coset leader of the corresponding coset, and decode the message.

```{r}
G <- matrix(c(1,0,0,0,0,1,1,1,0,1,1,0,0,1,0,
              0,1,0,0,0,0,1,1,1,0,1,1,0,0,1,
              0,0,1,0,0,1,1,0,1,0,1,1,1,1,0,
              0,0,0,1,0,0,1,1,0,1,0,1,1,1,1,
              0,0,0,0,1,1,1,0,1,1,0,0,1,0,1),
            byrow=TRUE, nrow=5)
mess <- c(0,0,1,0,1,1,1,1,1,1,0,0,0,0,1)
```

