---
title: "Systematic Codes"
output:
  html_document:
    toc: false
    df_print: paged
    theme: spacelab
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(comment = NA)
```

Upload to Canvas a PDF of your work for the following problems. 

## Assignment: due Monday, 11:59 pm

1. Use the fact that $vH^T = \mathbf{0}$ for any codeword $v$ to prove both parts of the following lemma.

> **Lemma.** Let $C$ be a systematic linear binary code, and let $H$ be its parity check matrix.
>
> 1. If $C$ contains a codeword of weight 1, then $H$ contains a zero column.
> 2. If $C$ contains a codeword of weight 2, then $H$ has two identical columns.

2. The following questions refer to the generating matrix `G` and message `mess` given below. 
    a. The matrix `G` is a generating matrix for a systematic linear binary code $C$. How many codewords are there in this code?
    b. Generate all the codewords in $C$, and find $d(C)$ by finding the weight of the lightest nonzero codeword in $C$.
    c. How many errors can this code correct?
    d. How many different syndromes are there?
    e. Compute the parity check matrix $H$.
    f. Bob receives the message `mess`. Find its syndrome.
    g. Determine the coset leader of the coset of `mess`, and decode the message.

```{r}
G <- matrix(c(1,0,0,0,0,1,1,1,0,1,1,0,0,1,0,
              0,1,0,0,0,0,1,1,1,0,1,1,0,0,1,
              0,0,1,0,0,1,1,0,1,0,1,1,1,1,0,
              0,0,0,1,0,0,1,1,0,1,0,1,1,1,1,
              0,0,0,0,1,1,1,0,1,1,0,0,1,0,1),
            byrow=TRUE, nrow=5)
mess <- c(0,0,1,0,1,1,1,1,1,1,0,0,0,0,1)
```

