---
title: "Cyclic Codes"
output:
  html_document:
    toc: false
    df_print: paged
    theme: spacelab
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(comment = NA)
```

Upload to Canvas a PDF of your work for the following problems. 

## Assignment: due Monday 11/30, 11:59 pm

1. Hamming codes have the following properties:
    - The nonzero syndromes are the columns of the parity check matrix $H$.
    - For nearest neighbor decoding, the column of the syndrome is the location of the error.
    
The following commands compute the parity check matrix for the Hamming $[63,57]$ code. 

```{r}
m <- 6
powersOf2 <- 2^(0:(m-1))
ham63H <- sapply(c(setdiff(1:(2^m-1), powersOf2), powersOf2), 
                 function(x){as.numeric(intToBits(x)[1:m])})
```

The following message contains a single error. Decode it (without computing the entire coset and without a coset leader syndrome table). In which position was the error? 

```{r}
mess <- c(1,1,1,0,0,0,0,0,1,1,0,0,1,0,1,0,1,1,1,0,0,1,0,1,0,0,0,0,0,1,0,1,1,1,0,1,0,1,1,1,0,1,1,0,1,0,1,0,0,0,1,1,0,1,0,0,1,1,1,0,1,0,0)
```

2. Let $C$ be the (unique) $[15, 12]$ code residing in $\mathbb{Z}_2/(x^{15} + 1)$.
    a. Use [Macaulay2](http://web.macaulay2.com/) to factor $x^{15}+1$.
    b. Find the generating polynomial for $C$.
    c. Find a parity check polynomial for $C$.
    d. Determine whether $x^{14} + x^{9} + x^5 + x^4 + x^2 + x + 1$ is a code word. Show how you can tell.

3.  Let $C = \langle x^4+x^3+x^2+x+1 \rangle$ in $\mathbb{Z}_2/(x^{2055} + 1)$.
    a. How many code words does $C$ have?
    b. Show that the minimum distance of $C$ is less than or equal to 2.
    c. Show that $C$ contains a code word of weight $20$. Find such a code word, and explain how you know it is a code word.
