---
title: "The McEliece Cryptosytem"
output:
  html_document:
    toc: false
    df_print: paged
    theme: spacelab
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(comment = NA)
```

Upload to Canvas a PDF of your work for the following problems. You will definitely want to type this one. A text listing of your R commands and output will suffice. 

## Assignment: due Wednesday, 11:59 pm

The Golay code $G_{24}$ is a linear $[24,12]$ code that can correct $t=3$ errors. The following lines of code will load the generating matrix `golay24` and coset leader syndrome table `gT` into your local R environment.

```{r, eval=FALSE}
golay24 <- matrix(c(1,0,0,0,0,0,0,0,0,0,0,0,1,1,1,0,1,1,1,0,0,0,1,0,
                    0,1,0,0,0,0,0,0,0,0,0,0,1,0,1,1,0,1,1,1,0,0,0,1,
                    0,0,1,0,0,0,0,0,0,0,0,0,1,1,0,1,1,0,1,1,1,0,0,0,
                    0,0,0,1,0,0,0,0,0,0,0,0,1,0,1,0,1,1,0,1,1,1,0,0,
                    0,0,0,0,1,0,0,0,0,0,0,0,1,0,0,1,0,1,1,0,1,1,1,0,
                    0,0,0,0,0,1,0,0,0,0,0,0,1,0,0,0,1,0,1,1,0,1,1,1,
                    0,0,0,0,0,0,1,0,0,0,0,0,1,1,0,0,0,1,0,1,1,0,1,1,
                    0,0,0,0,0,0,0,1,0,0,0,0,1,1,1,0,0,0,1,0,1,1,0,1,
                    0,0,0,0,0,0,0,0,1,0,0,0,1,1,1,1,0,0,0,1,0,1,1,0,
                    0,0,0,0,0,0,0,0,0,1,0,0,1,0,1,1,1,0,0,0,1,0,1,1,
                    0,0,0,0,0,0,0,0,0,0,1,0,1,1,0,1,1,1,0,0,0,1,0,1,
                    0,0,0,0,0,0,0,0,0,0,0,1,0,1,1,1,1,1,1,1,1,1,1,1),
                  byrow=TRUE, nrow=12)
load(url("http://math.westmont.edu/data/golay24table.RData"))
gT[["001101011010"]] # example: use a syndrome to look up the coset leader
```

1. Use the functions `randInvMatrix` and `randPermMatrix` to make Bob's choice of appropriately-sized $S$ and $P$, respectively. Use at least 200 operations when you randomize. **Tip:** Set the random number seed using `set.seed` so you will be able to reproduce your work. Give $S$, $P$, and also the public key $G_1$ that Bob publishes.

2. Alice wishes to send the 12-bit message `101010101010`. Use the `randBinVector` function to create an appropriately-sized and weighted random error vector $e$, and compute the binary vector $y$ that Alice sends. 

3. Show how Bob can decrypt the message by computing $y_1$, computing its syndrome, using the coset leader syndrome table `gT` to obtain the code word $x_1$, and then applying $S^{-1}$ to the information bits to recover Alice's original 12-bit message.

## Helper Functions

```{r eval=FALSE}
randInvMatrix <- function(k, numops = 100) {
  M <- diag(k) # the k by k identity matrix
  Minv <- diag(k)
  rowOps <- replicate(numops, sample(1:k, 2))
  for(i in 1:numops) {
    M[rowOps[1,i],] <- (M[rowOps[1,i],] + M[rowOps[2,i],]) %% 2
    Minv[rowOps[1,numops-i+1],] <- (Minv[rowOps[1,numops-i+1],] + Minv[rowOps[2,numops-i+1],]) %% 2
  }
  return(list(M = M, Minv = Minv))
}
```

```{r eval=FALSE}
randPermMatrix <- function(n, numswaps = 100) {
  M <- diag(n)
  rowSwaps <- replicate(numswaps, sample(1:n, 2))
  for(i in 1:numswaps) {
    M[rowSwaps[,i],] <- M[c(rowSwaps[2,i],rowSwaps[1,i]),]
  }
  return(M)
}
```

```{r eval=FALSE}
randBinVector <- function(length, weight) {
  v <- numeric(length)
  oneLocs <- sample(1:length, weight)
  v[oneLocs] <- 1
  return(v)
}
```